### 2018-04-02 paper.tex
Update on Overleaf.

\documentclass[letterpaper]{article}

\usepackage{natbib,alifeconf}  %% The order is important

% *****************
%  TODOs:
% *****************
% [ ] Writing: Abstract
% [ ] Writing: Results & discussion
% [ ] Writing: Conclusions
% [ ] Big 'ol citation pass: need to work appropriate citations into paper

% *****************
%  Requirements:
% *****************
%
% - All pages sized consistently at 8.5 x 11 inches (US letter size).
% - PDF length <= 8 pages for full papers, <=2 pages for extended
%    abstracts.
% - Abstract length <= 250 words.
% - No visible crop marks.
% - Images at no greater than 300 dpi, scaled at 100%.
% - Embedded open type fonts only.
% - All layers flattened.
% - No attachments.
% - All desired links active in the files.

% Note that the PDF file must not exceed 5 MB if it is to be indexed
% by Google Scholar. Additional information about Google Scholar
% can be found here:
% http://www.google.com/intl/en/scholar/inclusion.html.

% If your system does not generate letter format documents by default,
% you can use the following workflow:
% latex example
% bibtex example
% latex example ; latex example
% dvips -o example.ps -t letterSize example.dvi
% ps2pdf example.ps example.pdf

% For pdflatex users:
% The alifeconf style file loads the "graphicx" package, and
% this may lead some users of pdflatex to experience problems.
% These can be fixed by editing the alifeconf.sty file to specify:
% \usepackage[pdftex]{graphicx}
%   instead of
% \usepackage{graphicx}.
% The PDF output generated by pdflatex should match the required
% specifications and obviously the dvips and ps2pdf steps become
% unnecessary.

% Note:  Some laser printers have a serious problem printing TeX
% output. The use of ps type I fonts should avoid this problem.

\title{Understanding fine temporal scale evolutionary dynamics with lineage metrics}
\author{Emily Dolson$^{1,2,3}$, Alexander Lalejini$^{1,2, 3}$, Steven Jorgensen$^{1,2}$ \and Charles Ofria$^{1,2, 3}$ \\
\mbox{}\\
$^1$BEACON Center for the Study of Evolution in Action, Michigan State University, East Lansing, MI, 48824 \\
$^2$Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, 48824 \\
$^3$Ecology, Evolutionary Biology, and Behavior Program, Michigan State University, East Lansing, MI, 48824 \\
dolsonem@msu.edu} % email of corresponding author

% For several authors from the same institution use the same number to
% refer to one address.
%
% If the names do not fit well on one line use
%         Author 1, Author 2 ... \\ {\Large\bf Author n} ...\\ ...
%
% If the title and author information do not fit in the area
% allocated, place \setlength\titlebox{<new height>} after the
% \documentclass line where <new height> is 2.25in

\begin{document}
\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

Evolution is an emergent effect of events (\textit{e.g.} replication, variation, and competition) that occur on a much finer-grained temporal scale. While evolution's emergent nature is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results. In computational evolutionary systems, we theoretically have adequate data to untangle these mechanisms. In practice, though, gathering appropriate data to do so can be challenging. Moreover, because of the vast difference between the scales that evolutionary causes and effects can occur on, the quantity of such data can be overwhelming. Both of these problems can be mitigated with a standardized set of diagnostic summary metrics that paint a picture of shorter-term evolutionary dynamics within a population. Here, we suggest a suite of metrics that operate on lineages and phylogenies. We then use experimental results to provide an intuition for what they can tell us about evolution.

A lineage describes a continuous line of descent, linking ancestors and their descendants. A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation. Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in modern wet lab experiments, allowing us to replay the tape of life in exacting detail and to tease apart the evolutionary recipe for whatever complex feature or phenomenon we happen to be interested in~\citep{mcphee_using_2016}. 
In a more famous example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}.

Yet, tracking the full details of a single lineage, much less a population of lineages, can be computationally expensive and will inevitably generate an unwieldy amount of data that can be challenging to visualize and interpret~\citep{mcphee_visualizing_2016}.
Summary statistics can help alleviate these issues by enabling the user to focus on aggregate trends across a population rather than needing to examine each individual's lineage. The question is how to effectively summarize what is essentially a path through fitness space. A useful abstraction is to treat the path as a sequence of states. Here, we will use phenotypes and genotypes as the states in the sequence, but we could just as easily use some other descriptor of the lineage's position in the fitness landscape at a given point in time.
With this abstraction in hand, a few metrics are easily formalized: the number of unique states, the number of transitions between states, and the amount of time spent in each state. 
Additionally, we may care about how the transitions between states happened. What mutations lead to them? Were those mutations beneficial, deleterious, or neutral at the time? Here, we choose a subset of these metrics that we expect will be particularly useful to explore further.

Whereas a lineage represents the evolutionary history of a single individual, a phylogeny represents the evolutionary history of an entire population. Measurements that summarize phylogenies can provide useful insight into population-level evolutionary dynamics, such as diversification and co-existence among different clades. A variety of useful phylogeny measurements have already been developed by biologists~\citep{tucker_guide_2017}. These measurements tend to treat the phylogeny as a graph and make calculations about its topology. Tucker \textit{et al.} group them into three broad categories: assessments of the quantity of evolutionary history represented by a population, assessments of the amount of divergence within that evolutionary history, and assessments of the topological regularity of the phylogenetic tree. Such measurements can help quantify the behavior of the population as a whole, providing insight into interactions between its members. Thus, they are useful indicators of the presence of various types of eco-evolutionary dynamics.
%@AML: I think the next bit (just moved it) might belong in the metrics section. Instead, we might want a little more elaboration on each of the categories of metrics.
%@ELD: Is that too handwavy?
%@AML: Nah! Handwaving is just what we needed!

Here, we present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility. Additionally, we suggest the adoption of four phylogeny measurements from biology: most-recent common ancestor depth, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity. For each metric, we discuss its application and our intuitions for what it can tell us about evolution. We evaluate our intuitions on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths. For this work, we restrict our attention to asexually reproducing populations; however, we make suggestions as to how these metrics can be extended to sexual populations.

%@AML:  It might be nice to wrap this section up with a little bit of a 'call to action'/'why should you care about this as a community'
In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we hope this work begins a conversation within the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories. There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are incredibly valuable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history. However, we are unaware of efforts to formalize a suite of quantitative lineage- and phylogeny-related metrics for computation evolution. Indeed, .... % people have done stuff with lineages
% - often a component of a larger experiment-specific analysis or limited to qualitative descriptions/visualizations
% - well-defined metrics not only provide valuable tools for teasing apart evolutionary dynamics, they also facilitate allow us to move away from exclusively qualitative descriptions/understandings of results ==> better quantitative understanding/statistics ==> better hypothesis testing
% - facilitate comparisons among systems. 
% - Something about activity metrics?

\section{Metrics}

Code for all of our metrics is open source and available in the Empirical library (https://github.com/devosoft/Empirical). Empirical is a C++ library built to facilitate writing efficient and easily sharable scientific software. To facilitate including useful features without excessive overhead, Empirical is header-only, meaning these metrics can be imported into an existing project with minimal effort.
% [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager?

\subsection{Lineage Metrics}
% Outlining some points to (maybe) make here; probably too bogged down in details; not even sure we want to talk about lineages as graphs:
% - Apply metrics in context of asexual populations
% - Establish a framework from which to think about metrics (not novel, but rarely formalized)
% 	 - We treat lineages as sequences of states where each state represents an individual and transitions between states represent parent-offspring relationships.
%      - More formally, lineages are directed acyclic graphs where internal nodes have in-degree = 1 (parent), out-degree = 1 (offspring). Original ancestor (in digital evolution, often seed of population) w/out-degree = 1. Extant individual w/in-degree = 1.
%      - Transitions specify relevant variation introduced into offspring (e.g. mutations)
%      - States specify relevant genotype and phenotype information (e.g. genome, location in space, behavior, etc) -- anything we might want to measure over a lineage
%      - Becomes more complicated with sexual populations, but we'll stick with asexual assumption for this. 
%  - Already, framework affords certain questions:
%     - How many individuals are along the lineage? 
%     - Total # and magnitude of mutations along lineage?
%     - Distribution of mutation types (ins vs. sub/beneficial vs. deleterious)?
%  - Further abstractions:
%      - Sequence of genotypes along lineage (where each genotype may represent multiple individuals)
%      - Sequence of phenotypes along lineage (where each phenotype may represent multiple individuals)
%      - Still further abstraction: sequence of a particular aspect of genotype (e.g. genetic marker)/phenotype (e.g. location) along lineage
%  - We discuss three classes of lineage metrics within this framework: lineage length, mutation accumulation, and phenotypic volatility. 
%  - For each, we discuss our intuition for what they can tell us (alone and in combination) about evolution. 
Each of the three lineage metrics we discuss --- lineage length, mutation accumulation, and phenotypic volatility --- reduces a lineage to a linear sequence of states where each state represents an individual or a continuous sequence of individuals that share some common genotypic or phenotypic characteristic of interest. While we limit our focus to three lineage metrics, this abstraction places lineages in a form suitable for a wide range of measurements, including the direct application of many data mining techniques designed to operate over sequences such as sequential pattern mining, trend analysis, \textit{et cetera} \citep{han2011data}.

Only lineages consisting of asexually reproducing ancestors where genetic material is exclusively vertically transmitted can be directly abstracted as a \textit{linear sequence} of states. 
Sexual reproduction (and any form of horizontal gene transfer) complicates matters significantly as such lineages are more appropriately represented by trees rooted at the extant organism, branching for each contributor of genetic material. 
For each metric we present here, we limit our discussion to asexual populations; however, we suggest two approaches for generalizing these lineage metrics to sexual populations: apply a lossy compression to reduce sexual lineages into linear sequences or extend our application of the metrics to operate over non-linear (tree) structured lineages. 
Sexual lineages can be compressed into linear sequences of states by modeling sexual reproduction events as asexual reproduction events, designating one parent to be a part of the lineage and considering the genetic contributions of other parents as sources of genetic variation (mutations). The primary downside to this approach is its lossy-ness (i.e. the fact that it discards potentially important parentage information). 
Alternatively, we can extend our metrics to operate over the more complex state sequences that constitute the lineages of sexually reproducing organisms. One such approach would be to consider all possible ancestor paths for an extant individual, calculating a given metric for each of them and then averaging the resulting values together. Assessing the efficacy of these and potentially other approaches would be a useful line of research to pursue in the future.

\subsubsection{Lineage Length}
% Define/explanation
% 	- Within context of framework, lineage length is simply measure of number of states in the graph/sequence.
%   - What this means will depend on what states represent?
%       - number of individuals along lineage 
%             - More interesting in experiments where generations are not dictated by system, but rather by life history strategies of organisms
%             - Give you a diagnostic for replication rate (r vs. k strategy)
%      - number of genotypes along lineage
%            - length = 1: no mutations
%            - length = number of individuals: mutation in every replication
%            - Intuitively, should be higher at high mutation rates, lower at low mutation rates. 
%      - number of phenotypes along lineage
%            - low: evidence for static environment
%            - high: might be indicative of a changing environment or ecological interactions
%            - if number of genotypes is high, but number of phenotypes low, probably in a neutral fitness landscape
%      - various subsets of genotype/phenotype
%           - location in space (environment type): useful for looking at lineage through heterogeneous landscape
%           - change in particular phenotypic trait
%           - change in genetic markers
%           - Distributions of above two lengths can tell you useful things about drift, selection, etc.

Lineage length describes the number of ancestral states traversed by a lineage. If a state represents a single individual along a lineage, lineage length is a measure of the number of generations on a lineage. This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined but instead determined by the life history strategies of organisms. In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates and shorter lineages imply slower replication rates. 
%@ELD: I've always heard these called r- and k- selected rather than replicator, but I'm totally willing to believe these terms are used too.
%@AML: Yup. I have no idea if I've ever heard them referred to as 'r-replicator vs. k-replicator'. I'm going to get rid of the two parentheticals for now.

Lineage length becomes an increasingly flexible and informative metric if we begin to consider more abstract definitions of an ancestral state along a lineage. 
We might measure lineage length where an ancestral state represents a temporally continuous sequence of individuals that share a phenotypic or genotypic characteristic of interest. In these cases, lineage length only increases if the characteristic of interest changes from parent to offspring. For example, in an environment where organisms must perform a set of tasks to be successful, we might define ancestral state to be the set of tasks performed by an individual (\textit{e.g.} the set of logic operations performed by a digital organism in the default logic 9 Avida environment). In this scenario, lineage length would only increase when the set of tasks performed by an ancestor changes, compressing sequential ancestors that perform the identical sets of tasks into a single ancestral state. 
% This could be made very clear with a good example here. 
%@ELD Maybe I'm just getting tired, but I didn't quite follow that last sentence
%@AML: It was definitely hard to follow. Hopefully it's better now? 
%  -- Still could use a sentence about what the measure might tell us in the given example.

\subsubsection{Mutation Accumulation}
Mutation Accumulation defines a set of measurements that track mutational changes across a lineage. These changes can be measured in a variety of ways, such as the magnitude of the change (for real-valued genomes), or the total count of changes (for discrete-valued genomes). The type of a mutation can also be tracked, and can provide insight into the distribution of mutation types for a given lineage. Measures of mutation accumulation along the lineages of successful extant individuals can help to tease apart the relative importance of different types of mutational events when compared to what is expected by chance.

In conjunction with collected fitness information, the class of a mutation (\textit{e.g.} beneficial, deleterious, or neutral) can also be tracked. Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.   
The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys. 
Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.

% Mutation accumulation over a lineage 
%  - Measured in a variety of ways
%   - For real-valued genomes, we can track magnitudes of changes
%   - For discrete-valued genomes, we can track counts of changes
%      - Track by type to get distribution of mutation types
%   - In conjunction with fitness information, can track classes of mutation: beneficial, deleterious, neutral
%      - Again, can extract a distribution; different environments/experiments, you might expect different distributions
%      - Importance of deleterious steps; valley crossing, etc

\subsubsection{Phenotypic Volatility}

Phenotypic volatility addresses the rate at which the lineage changes phenotype (although the same concept can be applied to other types of state). In systems with discrete/categorical phenotypes, this can be measured by summing the number of times the phenotype changes. A related but subtly different measurement in such systems is the number of unique phenotypes on a lineage. In most cases, these values will be very similar. A discrepancy between them, however, would be a clear signal of interesting dynamics. Specifically, it would suggest that the lineage was in a portion of the landscape where it could rapidly switch among different phenotypes. Such behavior could be indicative of some form of evolutionary bet-hedging.

In systems with continuous-valued phenotypes, a subtly different approach is needed to measure phenotypic volatility, because there are no discrete state transitions. Instead, we can measure the overall variance in phenotype along a lineage. In some cases, it may be desirable to smooth out the noise inherent in a real-valued phenotype. We can do so by instead taking the variance of the rolling mean of fitness, to more closely approximate the idea of measuring phase transitions.
% TODO

% Define/describe
% .... 
% Useful things that this metric could identify:
% - Mutational bet hedging (high volatility)
% - Plasticity (high volatility)
% - Neutral landscapes (expect low volatility)

% Here, we...

\subsubsection{Summary statistics}

Each of these metrics can be calculated for each member of the population at each time step. Doing so, however, would produce an amount of data so large that it would be difficult to make sense of. Instead, we need to come up with ways to generate useful summaries. There are two main approaches to doing so: 1) choose a small number of representative lineages from a given time point, 2) collect summary statistics about the distribution of metric values across the population.

A single representative lineage can be chosen by selecting the lineage of the dominant organism (either the most fit or the more numerous; here we use the most fit). In populations with a diversity of coexisting strategies this approach can be uninformative as the dominant lineage is not necessarily representative of all successful lineages. An alternative in these scenarios is to filter out lineages that do not have offspring some predetermined number of generations later as such lineages were likely not representative of an important subset of the population. Still, any approach based on measuring only a subset of lineages can be challenging to interpret when the current dominant lineage (or lineages) is replaced with a different one; such changes can introduce a discontinuity if the value is being measured over time. If graceful responses to changes in which lineage is dominant are required, it can be advantageous to instead measure summary statistics (e.g. mean, variance, and range) across the entire population.

In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be very similar to the average lineage in the population, as most of the population will be closely related. When the population contains more phylogenetic diversity, however, it is possible for the dominant lineage to differ from the mean. The nature of such differences is likely informative about the evolutionary dynamics occurring in the population. Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population.

\subsection{Phylogeny metrics}

These metrics operate on entire phylogenies rather than single lineages within a population. Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics. Because they factor in the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage. On the other hand, because most lineages in the population share a substantial amount of history, phylogeny metrics can often be calculated more rapidly than lineage metrics can be calculated across an entire population. Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc. Thus, when we refer to items in a phylogeny, we will use the term \textit{taxa}.

A standard technique for saving memory and time when working with lineages in computational systems is to "prune" them, removing any taxa that are not either currently alive or the ancestor of taxa that are alive. Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies. Thus, for the remainder of this paper, we will assume we are only working with pruned phylogenies.

In populations without ecological forces promoting coexistence, phylogenies should coalesce periodically, resulting in pruned lineages that mostly consist of a single path. When there is strong selection, this coalescence should happen even more rapidly. Thus, phylogenies with topologies that deviate from that expectation are an indication of ecological interactions within the population. The metrics discussed here can provide insight into the nature of those interactions and their long-term evolutionary effects. As a result, they are often referred to as phylogenetic diversity metrics~\citep{tucker_guide_2017}.

An important distinction between phylogenies in natural vs computational systems to be aware of is that in natural systems phylogenies are generally inferred from extant taxa, whereas in computational systems they are directly measured. The inferred phylogenies used in natural systems do not contain internal nodes unless those nodes are branch points. They also do not contain history prior to the most recent common ancestor (MRCA) of all organisms in the extant population. For consistency, we will exclude pre-MRCA taxa from our analyses. However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.

Here we provide a high-level summary of phylogeny metrics that we expect will be particularly useful. For more metrics and more detail on all of these metrics, see~\citep{winter_phylogenetic_2013, tucker_guide_2017}.

\subsubsection{Depth of Most-Recent Common Ancestor}

The depth of the MRCA (i.e. how many steps from the beginning of time did it occur?) is a useful metric because of the relatively high ratio of informativeness to ease of calculation. A more recent MRCA implies more frequent selective sweeps and less long-term stable coexistence between clades. The depth of the MRCA is so easily calculated because most any implementation of phylogeny pruning requires keeping track of it already. Measuring the frequency with which the MRCA changes (i.e. the number of coalescence events) can also by informative, as some conditions can inflate the length of the lineage relative to other conditions, without actually increasing the frequency of selective sweeps. This scenario is particularly likely when the population size is changing over time. A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never see a change in the MRCA after the very beginning of evolution.

\subsubsection{Phylogenetic Richness}

Measurements of phylogenetic richness quantify the total amount of evolutionary history that is contained in a set of taxa. The most traditional metric of phylogenetic richness is the one simply titled "Phylogenetic Diversity", which is calculated as the number of the nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}. Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}. A third approach is to calculate evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, and sum it across all extant taxa~\citep{tucker_guide_2017}.

\subsubsection{Phylogenetic Divergence}

Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other, and are often the means of values that can be calculated for individual taxa. For instance, one option is to calculate the mean of the pairwise distances between all taxa in the population~\citep{webb_exploring_2000}. Similarly, phylogenetic divergence can also be calculated by calculating the evolutionary distinctiveness of each taxon in the population (as alluded to in the previous section) and taking the average of these values.

\subsubsection{Phylogenetic Regularity}

Measurements of phylogenetic regularity quantify how balanced the branches of the phylogeny are, and are often the variances of values that can be calculated for individual taxa. Just as the mean of the pairwise distances between all taxa in the population is a measurement of phylogenetic divergence, taking their variance is a measurement of phylogenetic regularity. The same is true of the variance of evolutionary distinctiveness across the population.

\section{Test Problems}
To understand the metrics defined above, the test problems used need to be well understood and studied. The benchmark functions from the GECCO Competition on Niching Methods meet both of these requirements and allow us to visualize the actual fitness landscape, due to the low dimensionality of the problems ~\citep{li_benchmark_2013}. For each problem, the X and Y coordinates offered by a given organism are translated by the function into a fitness value. We chose a diverse subset of these functions (Himmelblau, Shubert, Composition Function 2, and Six-Humped Camel Back) as our test problems in order to gain a broad understanding of our metrics. We used the implementations of these problems at https://github.com/mikeagn/CEC2013/ (C++ for fitness calculations during evolution, Python for post-hoc analysis). Figure \ref{TODO} shows the fitness landscapes defined by each of our four chosen test problems.

% @AML: It would be neat to have a tournament size = population size treatment (pure hill climbing)
For each test problem, we evolved populations of 1000 digital organisms under a range of mutation rates and selection strengths for 5000 generations. Each digital organism's genome consisted of two floating point numbers that defined its position in the fitness landscape. To determine which organisms reproduced each generation, we used tournament selection. We evolved populations under five different tournament sizes: one, two, four, eight, and sixteen.  Tournament size represents strength of selection where higher tournament sizes correspond to a higher strength of selection and lower tournament sizes correspond to a lower strength of selection. A tournament size of one is equivalent to no selection pressure (\textit{i.e.} every organism in the population has an equal chance of being selected to reproduce). Organisms selected to reproduce did so asexually. Each value in an offspring's genome was mutated by adding noise given by a normal distribution with a mean of 0; the `mutation rate' of a treatment defined the standard deviation used to defined this normal distribution and was given as a proportion of the test problem's domain.  For each problem and tournament size, we evolved populations at eight mutation rates: \textbf{TODO}. 
% @AML: Don't love how I've phrased mutation rate sentences.

% @AML: if room, it might be helpful to have a cartoon figure that gives all of our treatments
%		-- I'm imagining a table where we have axes/arrows that show 'increasing selection' vs. 'increasing mutation rate'

\section{Implementation}

\section{Data Analysis}
\subsection{3D visualizations}
% Trim dramatically
%Evolutionary biologists often accuse each other of making up "just-so" stories - post-hoc explanations for how an observed trait or pattern could have evolved. The problem with these stories is that, in biology, there is often no way to verify that they reflect what happened in biology. Due to evolution's stochastic nature, it is easy to come up with possible ways that something could have happened once, regardless of how repeatable they are. Theoretically, we should easily be able to avoid this problem in computational evolution by carefully checking the underlying assumptions behind our hypothesized explanation.  All too often, however, we fail to drill down to the true underlying mechanism behind an observed effect.

%The metrics we propose here can help provide a check on this behavior, by making information about underlying evolutionary dynamics more readily accessible. They can only have that effect, however, if we completely understand what they are telling us about the way populations under different conditions are traversing different fitness landscapes. In order for these metrics to be useful for diagnosing the behavior of an evolving population, we need to establish ground truth for what underlying evolutionary dynamics are implied by different values of the metrics. For the purposes of building this intuition on a solid foundation, we wanted to be able to visualize the full evolutionary history of each population as it traversed the fitness landscape.

%To this end, we chose problems for which we could perfectly visualize the fitness landscape, and kept track of the complete evolutionary history of all members of the population. With these data, we were able to overlay successful lineages on top of the fitness landscape. Visualizing all lineages in each experimental condition in this way gave us an efficient check on our understanding of the underlying evolutionary dynamics in that condition. 
In order to make these metrics useful in the long run, it is critical that we have an accurate understanding of how various measurements correspond to the actual behavior of lineages. The most direct way to confirm our expectations is to visualize the path that each lineage takes across the fitness landscape, mapping the x, y, and z (fitness) coordinates of each ancestor of each member of the population. Creating such a visualization entails incorporating a large quantity of information into a limited space. When projected onto two dimensions, lineages can obscure parts of the fitness landscape (and each other). To mitigate this problem, we built a thee-dimesnional visualization using the A-Frame Framework [cite].

A-Frame supports building three dimensional scenes and rendering them to a variety of platforms. In the simplest case, the visualization is rendered in WebGL, allowing it to be viewed in a standard web browser. Mouse interactions such as rotation make it possible to view the visualization from all angles, and WebGL's use of the graphics card allows it to render data-rich visualizations. A-Frame also supports rendering the page with WebVR, allowing it to be viewed using various virtual reality headsets. These platforms allow the user to explore the data in a truly three-dimensional environment. Different headsets support different amounts of interaction. For the data interpretation in this paper, we used an Oculus Rift to provide us with fine-grained control of which part of the visualization we were looking at.

\subsection{Metric analysis}

We analyzed trends in our metrics using the R Statistical Computing Language~\citep{r_core_team_r:_2017}. Specifically, we used the ggplot2 library for all graphs included in this paper~\citep{wickham_ggplot2:_2009}. All analysis scripts are available at https://github.com/stevenjson/ALife2018-Lineage.

\section{Results and Discussion}
% Changing mutation: here's what happens when we change mutation on each problem
\begin{figure*}
\includegraphics[width=7in]{figs/all_mutation_rate.png}
\caption{Values of example metrics across different mutation rates for each of the four problems. Circles are medians and vertical lines show inter-quartile range.}
\end{figure*}

% \begin{figure}
% \includegraphics[width=3.5in]{figs/mrca_depth_mutation.png}
% \end{figure}

% \begin{figure}
% \includegraphics[width=3.5in]{figs/variance_deleterious_mutation_rate.png}
% \end{figure}

% - Metric 1
% Problem 1
% Problem 2
% Problem 3
% Problem 4
% - Metric 2
% ...

\begin{figure*}
\includegraphics[width=7in]{figs/all_ts.png}
\caption{Values of example metrics across different tournament sizes for each of the four problems. Circles are medians and vertical lines show inter-quartile range.}
\end{figure*}
% \begin{figure}
% \includegraphics[width=3.5in]{figs/mrca_ts.png}
% \end{figure}

% Changing selection
% - Metric 1
% Problem 1
% Problem 2
% Problem 3
% Problem 4
% - Metric 2

\section{Conclusions}

\section{Acknowledgements}

\footnotesize
\bibliographystyle{apalike}
\bibliography{Zotero,bibliography} % replace by the name of your .bib file

\end{document}

### 2018-04-02 paper.tex
Update on Overleaf.

@@
 Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.
-The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys.
+The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys~\citep{.
 Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.

### 2018-04-02 paper.tex
Update on Overleaf.

--- 2018-04-02 paper.tex
@@
 Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.
-The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys.
+The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys~\citep{}.
 Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.

### 2018-04-02 analysis/data_analysis.Rmd
Added phenotypic volatility

---
title: "Lineage Metrics Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(readr)
library(reshape2)
library(dplyr)
library(scales)
```

```{r, echo=FALSE}
alex_real_value_data <- read_csv("alex_real_value_data.csv")
emily_real_value_data <- read_csv("real_value_data.csv")
real_value_data <- merge(alex_real_value_data, emily_real_value_data, all = TRUE)
real_value_data$problem <- sapply(as.character(real_value_data$p),switch,"0"="Himmelblau","1"="Six-hump camel back","2"="Shubert", "6"="Composition function 2")
real_value_data$problem <- as.factor(real_value_data$problem)
real_value_data$selection <- sapply(as.character(real_value_data$selection),switch,"0"="Tournament","1"="Lexicase","2"="Eco-EA", "3"="MAP-Elites", "4"="Roulette", "5"="Drift")
real_value_data$selection <- as.factor(real_value_data$selection)
real_value_data$seed <- as.factor(real_value_data$seed)
#real_value_data$ts <- as.factor(real_value_data$ts)
real_value_data$elite <- as.factor(real_value_data$elite)
#real_value_data$mutation_rate <- as.factor(real_value_data$mutation_rate)
tournament_data <- subset(real_value_data, real_value_data$selection == "Tournament" & real_value_data$elite == 0)
tournament_endpoints <- filter(tournament_data, update==5000)

all_dom_data <- read_csv("~/repos/ALife2018-Lineage/analysis/cec_python_library/all_dom_data.csv")
all_dom_data$update <- 5000
all_dom_data$seed <- all_dom_data$RANDOM_SEED
all_dom_data$RANDOM_SEED <- NULL
all_dom_data$PROBLEM <- NULL
all_dom_data$DATA_DIRECTORY <- NULL
all_dom_data$RUN_MODE <- NULL
all_dom_data$path <- NULL

tournament_endpoints <- merge(tournament_endpoints, all_dom_data)
melted_tournament_data <- melt(tournament_endpoints, id.vars = c("update", "selection", "seed", "elite", "mutation_rate", "le", "p", "problem", "g", "ts", "TOURNAMENT_SIZE", "FITNESS_INTERVAL", "RESOURCE_SELECT__OUTFLOW", "MUTATION_STD", "SELECTION_METHOD", "RESOURCE_SELECT__RES_AMOUNT", "HINT_GRID_RES", "RESOURCE_SELECT__RES_INFLOW", "LEXICASE_EPSILON", "SYSTEMATICS_INTERVAL", "RESOURCE_SELECT__FRAC", "RESOURCE_SELECT__COST", "RESOURCE_SELECT__MAX_BONUS", "POP_SNAPSHOT_INTERVAL", "POP_SIZE", "ELITE_SELECT__ELITE_CNT", "GENERATIONS"))
end_points_vars_of_interest <- melted_tournament_data %>%  filter(variable %in% c("mrca_depth", "dominant_deleterious_steps", "mean_evolutionary_distinctiveness", "variance_evolutionary_distinctiveness", "current_phylogenetic_diversity", "mean_pairwise_distance", "variance_pairwise_distance", "phenotypic_volatility"))

labels <- c(current_phylogenetic_diversity = "Phylogenetic diversity", dominant_deleterious_steps = "Dominant deleterious steps", mean_evolutionary_distinctiveness="Mean evo. distinctiveness", mean_pairwise_distance="Mean pairwise distance", mrca_depth="MRCA depth", variance_evolutionary_distinctiveness="Variance evo. distinctiveness", variance_pairwise_distance = "Variance pairwise distance", phenotypic_volatility="Phenotypic volatility")

```

```{r}
ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$mutation_rate==.001), aes(x=ts, y=value, color=problem, group=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="line", fun.y=median)+ theme_classic() + scale_x_log10("Tournament size", breaks=c(2,4,8,16)) + scale_y_continuous("", labels=comma) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels)) + scale_color_discrete("Problem") + theme(legend.position = "bottom")
ggsave("../figs/all_ts.png", width=7, height=5) 
```

```{r}
ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$ts==4), aes(x=mutation_rate, y=value, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="line", fun.y=median)+ theme_classic() + scale_x_log10("Mutation rate", labels=scientific) + scale_y_continuous("", labels=comma) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels)) +scale_color_discrete("Problem") +theme(legend.position = "bottom")
ggsave("../figs/all_mutation_rate.png", width=7, height=5)

```

### 2018-04-02 paper.tex
Update on Overleaf.

--- 2018-04-02 paper.tex
@@
 Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.
-The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys.
+% @AML: the below points likely have many citations that could
+The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys~\citep{covert_experiments_2013}.
 Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.

### 2018-04-03 paper.tex

--- 2018-04-02 paper.tex
@@
 \usepackage{natbib,alifeconf}  %% The order is important
+\usepackage{amsmath}
+\usepackage{hyperref}
 
@@
 
+% Alternative titles:
+%  - A Suite of Metrics for Understanding Evolutionary Dynamics/Histories
 \title{Understanding fine temporal scale evolutionary dynamics with lineage metrics}
@@
 \begin{abstract}
+We suggest a suite of diagnostic metrics that operate on lineages and phylogenies in digital evolution experiments with the aim of improving our capacity to quantitatively explore evolutionary histories in digital evolution experiments.
+We present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility. 
+Additionally, we suggest the adoption of four phylogeny measurements from biology: most-recent common ancestor depth, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity. 
+We demonstrate the use of each metric on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths, confirming our intuitions about what they can tell us about evolutionary dynamics.
 
@@
 
-Evolution is an emergent effect of events (\textit{e.g.} replication, variation, and competition) that occur on a much finer-grained temporal scale.
+Evolution is a collective effect of many smaller events such as replication, variation, and competition that occur on a much finer-grained temporal scale.
 While evolution's emergent nature is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results.
-In computational evolutionary systems, we theoretically have adequate data to untangle these mechanisms.
-In practice, though, gathering appropriate data to do so can be challenging.
-Moreover, because of the vast difference between the scales that evolutionary causes and effects can occur on, the quantity of such data can be overwhelming.
-Both of these problems can be mitigated with a standardized set of diagnostic summary metrics that paint a picture of shorter-term evolutionary dynamics within a population.
-Here, we suggest a suite of metrics that operate on lineages and phylogenies.
+In computational evolutionary systems, we can collect data to help untangle these mechanisms.
+In practice, though, gathering this data can be challenging.
+Moreover, because evolutionary effects can be witnessed on a vastly different scale from their causes, the quantity of such data can be overwhelming.
+Both of these problems can be mitigated with a standardized set of diagnostic metrics that summarize shorter-term evolutionary dynamics within a population.
+Here, we suggest a suite of such metrics that operate on lineages and phylogenies.
 We then use experimental results to provide an intuition for what they can tell us about evolution.
@@
 In a more famous example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}.
+%@SPJ: 'More famous example' sounds a bit weird to me
 
@@
 %@AML:  It might be nice to wrap this section up with a little bit of a 'call to action'/'why should you care about this as a community'
-In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we hope this work begins a conversation within the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories.
+%@ELD: I am in favor.
+In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we intend for this work to begin a conversation within the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories.
 There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are incredibly valuable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history.
-However, we are unaware of efforts to formalize a suite of quantitative lineage- and phylogeny-related metrics for computation evolution.
+However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-related metrics for computation evolution.
 Indeed, ....
@@
 % [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager? 
+%@ELD: They should just need the systematics manager. And they might need to hook up a couple of data collection signals.
 
@@
 This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined but instead determined by the life history strategies of organisms.
-In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates and shorter lineages imply slower replication rates.
+In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates (\textit{e.g.} r-selected lineage) and shorter lineages imply slower replication rates (\textit{e.g.} K-selected lineage).
 %@ELD: I've always heard these called r- and k- selected rather than replicator, but I'm totally willing to believe these terms are used too.
 %@AML: Yup. I have no idea if I've ever heard them referred to as 'r-replicator vs. k-replicator'. I'm going to get rid of the two parentheticals for now.
+%@ELD: I like the idea of putting it in that context. I was just suggesting we replace "replicator" with "selected"
+%@AML: Put the parantheticals back.
 
@@
 For example, in an environment where organisms must perform a set of tasks to be successful, we might define ancestral state to be the set of tasks performed by an individual (\textit{e.g.} the set of logic operations performed by a digital organism in the default logic 9 Avida environment).
-In this scenario, lineage length would only increase when the set of tasks performed by an ancestor changes, compressing sequential ancestors that perform the identical sets of tasks into a single ancestral state.
+In this scenario, lineage length would only increase when the set of tasks performed by an ancestor changes; any sequential ancestors that perform the identical sets of tasks would be compressed into a single state in the sequence.
 % This could be made very clear with a good example here. 
@@
 %@AML: It was definitely hard to follow. Hopefully it's better now?
+%@ELD: That definitely helped. I tried tweaking it a little more (I think the term "ancestral state" was confusing me, because I'm used to it being the state of the ancestor at the root of the tree)
 %  -- Still could use a sentence about what the measure might tell us in the given example.
@@
 \subsubsection{Mutation Accumulation}
-Mutation Accumulation defines a set of measurements that track mutational changes across a lineage.
+Mutation accumulation defines a set of measurements that track mutational changes across a lineage.
 These changes can be measured in a variety of ways, such as the magnitude of the change (for real-valued genomes), or the total count of changes (for discrete-valued genomes).
@@
 Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.
-The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys~\citep{.
+% @AML: the below points likely have many things that could be cited to support. 
+The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape, as well as how well a lineage was able to cross fitness valleys~\citep{covert_experiments_2013}.
 Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.
@@
 Specifically, it would suggest that the lineage was in a portion of the landscape where it could rapidly switch among different phenotypes.
-Such behavior could be indicative of some form of evolutionary bet-hedging.
+Such behavior could be indicative of some form of evolutionary bet-hedging~\citep{beaumont2009experimental}.
+% @AML: There are a whole lot of citations for bet-hedging..
+Just picked a popular Nature article on experimental evolution of bet hedging.
+Alternatively, could go with older paper suggesting bet-hedging as a possible strategy that exists.
+%@ELD: This one sounds good. Normally I'd say cite both, but we probably don't have the space.
 
@@
 We can do so by instead taking the variance of the rolling mean of fitness, to more closely approximate the idea of measuring phase transitions.
+% SPJ: What is meant by 'rolling mean' ?
+%@ELD: Would "moving average" be more familiar?
+
 % TODO
@@
 A single representative lineage can be chosen by selecting the lineage of the dominant organism (either the most fit or the more numerous; here we use the most fit).
-In populations with a diversity of coexisting strategies this approach can be uninformative as the dominant lineage is not necessarily representative of all successful lineages.
+In populations with a diversity of coexisting strategies, this approach can be uninformative as the dominant lineage is not necessarily representative of all successful lineages.
 An alternative in these scenarios is to filter out lineages that do not have offspring some predetermined number of generations later as such lineages were likely not representative of an important subset of the population.
@@
 The nature of such differences is likely informative about the evolutionary dynamics occurring in the population.
-Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper.
-Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population.
+%Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. 
+% @AML: Not sure what the above sentence is saying.
+%Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population.
+% @SPJ: I am not sure what argument this is referring to?
+The fact that there may or may not be a dominant lineage that differs from the mean?
+%@ELD: The point was just that sometimes the dominant lineage is different from the mean and that probably matters but we're not going to talk about it.
+Since we're going to be short on space, I think we can just cut these lines.
 
@@
 Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc.
-Thus, when we refer to items in a phylogeny, we will use the term \textit{taxa}.
+Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}.
+% @SPJ: could reword to 'Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}.',  might be more clear?
+%@ELD: sounds good!
 
-A standard technique for saving memory and time when working with lineages in computational systems is to "prune" them, removing any taxa that are not either currently alive or the ancestor of taxa that are alive.
+A standard technique for saving memory and time when working with lineages in computational systems is to ``prune" them, removing any taxa that are not either currently alive or the ancestor of taxa that are alive.
 Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies.
@@
 
-An important distinction between phylogenies in natural vs computational systems to be aware of is that in natural systems phylogenies are generally inferred from extant taxa, whereas in computational systems they are directly measured.
+An important distinction between phylogenies in natural versus computational systems to be aware of is that in natural systems phylogenies are generally inferred from extant taxa, whereas in computational systems they are directly measured.
 The inferred phylogenies used in natural systems do not contain internal nodes unless those nodes are branch points.
@@
 A more recent MRCA implies more frequent selective sweeps and less long-term stable coexistence between clades.
-The depth of the MRCA is so easily calculated because most any implementation of phylogeny pruning requires keeping track of it already.
+The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require keeping track of it already.
 Measuring the frequency with which the MRCA changes (i.e.
-the number of coalescence events) can also by informative, as some conditions can inflate the length of the lineage relative to other conditions, without actually increasing the frequency of selective sweeps.
+the number of coalescence events) can also be informative, as some conditions can inflate the length of the lineage relative to other conditions without actually increasing the frequency of selective sweeps.
 This scenario is particularly likely when the population size is changing over time.
-A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never see a change in the MRCA after the very beginning of evolution.
+A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never see a change in the MRCA after the very beginning of evolution (which is informative as a way of detecting stably coexisting branches in the phylogeny).
+% @SPJ: change to 'The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require ...' (i.e remove 'so' and 'most any')
+% @SPJ: The above sentence might need use some support (why is it informative?)
+%@ELD: Better?
 
@@
 Measurements of phylogenetic richness quantify the total amount of evolutionary history that is contained in a set of taxa.
-The most traditional metric of phylogenetic richness is the one simply titled "Phylogenetic Diversity", which is calculated as the number of the nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}.
+The most traditional metric of phylogenetic richness is the one simply titled ``Phylogenetic Diversity", which is calculated as the number of the nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}.
 Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}.
@@
 
-Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other, and are often the means of values that can be calculated for individual taxa.
+Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other and are often the means of values that can be calculated for individual taxa.
 For instance, one option is to calculate the mean of the pairwise distances between all taxa in the population~\citep{webb_exploring_2000}.
@@
 
-Measurements of phylogenetic regularity quantify how balanced the branches of the phylogeny are, and are often the variances of values that can be calculated for individual taxa.
+Measurements of phylogenetic regularity quantify how balanced the branches of the phylogeny are and are often the variances of values that can be calculated for individual taxa.
 Just as the mean of the pairwise distances between all taxa in the population is a measurement of phylogenetic divergence, taking their variance is a measurement of phylogenetic regularity.
@@
 \section{Test Problems}
+\begin{figure}
+\includegraphics[width=3.5in]{figs/landscapes.png}
+\caption{The fitness landscapes used in this experiment: A) Himmelblau, B) Six-humped Camel Back, C) Shubert, and D) Composition Function 2.
+Interactive versions available at \url{https://emilydolson.github.io/fitness_landscape_visualizations}.}
+\label{fig:landscapes}
+\end{figure}
+
 To understand the metrics defined above, the test problems used need to be well understood and studied.
@@
 We used the implementations of these problems at https://github.com/mikeagn/CEC2013/ (C++ for fitness calculations during evolution, Python for post-hoc analysis).
-Figure \ref{TODO} shows the fitness landscapes defined by each of our four chosen test problems.
+Figure \ref{fig:landscapes} shows the fitness landscapes defined by each of our four chosen test problems.
 
@@
 Each digital organism's genome consisted of two floating point numbers that defined its position in the fitness landscape.
+We initialized populations by randomly generating a number of organisms equal to the population size. 
 To determine which organisms reproduced each generation, we used tournament selection.
 We evolved populations under five different tournament sizes: one, two, four, eight, and sixteen.
-Tournament size represents strength of selection where higher tournament sizes correspond to a higher strength of selection and lower tournament sizes correspond to a lower strength of selection.
+Tournament size represents strength of selection where higher tournament sizes correspond to a higher strength of selection and lower tournament sizes correspond to a lower strength of selection~\citep{blickle1995mathematical}.
 A tournament size of one is equivalent to no selection pressure (\textit{i.e.} every organism in the population has an equal chance of being selected to reproduce).
@@
 Each value in an offspring's genome was mutated by adding noise given by a normal distribution with a mean of 0; the `mutation rate' of a treatment defined the standard deviation used to defined this normal distribution and was given as a proportion of the test problem's domain.
-For each problem and tournament size, we evolved populations at eight mutation rates: \textbf{TODO}.
-% @AML: Don't love how I've phrased mutation rate sentences.
+We prevented mutations from causing a value from exceeding the valid domain of the given problem. 
+For each problem and tournament size, we evolved populations at eight mutation rates: $1\text{e-}08$, $1\text{e-}07$, $1\text{e-}06$, $1\text{e-}05$, $1\text{e-}04$, $1\text{e-}03$, $1\text{e-}02$, and $1\text{e-}01$.
+% [.10, 0.01, 0.001, 0.0001, 0.00001,  0.000001, 0.0000001, 0.00000001]
+For all experimental conditions, we ran ten replicates, each with a unique random number seed. 
+Our experiment is implemented using the Empirical library; our implementation can be found at https://github.com/stevenjson/ALife2018-Lineage.
+% @AML: We might want to rename the repository to match whatever the title of this paper ends up being (e.g.
+ALIFE2018-TITLE).
 
@@
 %		-- I'm imagining a table where we have axes/arrows that show 'increasing selection' vs. 'increasing mutation rate'
-
-\section{Implementation}
 
@@
 When projected onto two dimensions, lineages can obscure parts of the fitness landscape (and each other).
-To mitigate this problem, we built a thee-dimesnional visualization using the A-Frame Framework [cite].
+To mitigate this problem, we built a thee-dimensional visualization using the A-Frame Framework [cite].
 
@@
 \section{Results and Discussion}
+
+% Figure X and Y show results @ details... 
+% All combinations can be found in supplementary material.
+
 % Changing mutation: here's what happens when we change mutation on each problem
@@
 \includegraphics[width=7in]{figs/all_mutation_rate.png}
-\caption{Values of example metrics across different mutation rates for each of the four problems. Circles are medians and vertical lines show inter-quartile range.}
+\caption{Values of example metrics across different mutation rates for each of the four problems.
+All experiments shown here used a tournament size of 4.
+Circles are medians and vertical lines show inter-quartile range.}
+\label{fig:mutation_rate}
 \end{figure*}
 
-% \begin{figure}
-% \includegraphics[width=3.5in]{figs/mrca_depth_mutation.png}
-% \end{figure}
-
-% \begin{figure}
-% \includegraphics[width=3.5in]{figs/variance_deleterious_mutation_rate.png}
-% \end{figure}
-
-% - Metric 1
-% Problem 1
-% Problem 2
-% Problem 3
-% Problem 4
-% - Metric 2
-% ...
+Overall, our results were consistent with what evolutionary theory would predict.
+As mutation rate increases, coalescence takes longer, as evidenced by the fact that the MRCA is farther back in time at higher mutation rates (see Figure \ref{fig:mutation_rate}).
+Consequently, phylogenetic richness (as measured by phylogenetic diversity) is higher at higher mutation rates.
+Phylogenetic divergence, measured here as mean pairwise distance between taxa, is similarly higher at higher mutation rates.
+Evolutionary distinctiveness, being another measurement of phylogenetic divergence, behaved almost identically (see Supplementary Material).
+Variance of evolutionary distinctiveness and pairwise distance between taxa (phylogenetic regularity metrics) behaved very similarly to the phylogenetic divergence metrics.
+This pattern makes sense, as most phylogenetic divergence on these landscapes will produce unbalanced phylogenetic trees.
+If there were stable coexistence between multiple clades, we would expect to see a reduced correlation between the phylogenetic divergence metrics and the phylogenetic regularity metrics.
+Increased mutation rate also increases the number of deleterious steps taken, a logical consequence of increasing the amount of mutation relative to the strength of selection.
 
@@
 \includegraphics[width=7in]{figs/all_ts.png}
-\caption{Values of example metrics across different tournament sizes for each of the four problems. Circles are medians and vertical lines show inter-quartile range.}
+\caption{Values of example metrics across different tournament sizes for each of the four problems.
+All experiments shown here used a mutation rate of .001.
+Circles are medians and vertical lines show inter-quartile range.}
+\label{fig:ts}
 \end{figure*}
-% \begin{figure}
-% \includegraphics[width=3.5in]{figs/mrca_ts.png}
-% \end{figure}
 
-% Changing selection
-% - Metric 1
-% Problem 1
-% Problem 2
-% Problem 3
-% Problem 4
-% - Metric 2
+Similarly, increasing tournament size generally increases the rate of coalescence, as higher tournament sizes correspond to stronger selection (see Figure \ref{fig:ts}).
+As a result, all of the measurements of phylogenetic richness and divergence decrease as tournament size increases.
+MRCA depth, on the other hand, increases, directly reflecting the increased frequency of selective sweeps.
+
+Surprisingly, there is no clear effect of tournament size on the count of deleterious steps along the dominant lineage.
+Values for all selection schemes and tournament sizes hover near 2,500, meaning that a deleterious step is taken in roughly half of the 5000 generations.
+This is partially an effect of mutation rate; at the lowest mutation rate, there is a clear trend towards fewer deleterious steps as tournament size increases (see Supplementary Material).
+However, the effect of mutation rate on the relationship between tournament size and dominant deleterious steps is complex, particularly for composition function 2 (see Supplementary Material).
+These trends likely share a common cause with the thresholding effect evident in Figure \ref{fig:mutation_rate}, where the number of deleterious steps along the dominant lineage abruptly climbs between mutation rates of 0.0000001 and .00001 and remains relatively flat over other mutation rates.
+Based on an inspection of the 3-dimensional fitness landscape visualizations, we can see that this is not an effect of lineages moving from peak-to-peak; at most mutation rates, they tend to remain on a single peak.
+Thus, we can infer that this effect is the result of a drift-like phenomenon where, at sufficiently high mutation rates, all members of the population are constantly somewhat displaced from the fitness peak.
 
@@
 
+% Recap!
+% In this work, we suggest a suite of lineage and phylogeny metrics aimed at improving our capacity to quantitatively understand observed evolutionary histories in digital evolution experiments.
+We applied these metrics to a set of toy problems, confirming theoretical expectations.
+
+The primary goal of this work is two-fold: 1) to communicate a set of useful metrics aimed at improving our capacity to quantitatively understand evolutionary histories in digital evolution experiments, and 2) to spark a conversation in the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories.
+With feedback from the community, we plan to expand the presented suite of lineage and phylogeny metrics, compiling accessible descriptions and examples of each metric.
+
+In this work we demonstrated the suite of lineage and phylogeny metrics on a set of toy problems where evolving digital organisms are overly simplistic.
+Next, we plan to apply these metrics to more complex digital organisms that we evolve in a variety of qualitatively different environments where we would expect to observe a wide range of evolutionary dynamics.
+
 \section{Acknowledgements}
+
+\setlength{\bibsep}{2pt}
 

### 2018-04-03 paper.tex

@@
 \begin{abstract}
-We suggest a suite of diagnostic metrics that operate on lineages and phylogenies in digital evolution experiments with the aim of improving our capacity to quantitatively explore evolutionary histories in digital evolution experiments.
+Fine-scale evolutionary dynamics can be challenging to tease out when focused on broad brush stokes of whole populations over long timespans.
+We propose a suite of diagnostic metrics that operate on lineages and phylogenies in digital evolution experiments with the aim of improving our capacity to quantitatively explore the nuances of evolutionary histories in digital evolution experiments.
 We present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility. 
-Additionally, we suggest the adoption of four phylogeny measurements from biology: most-recent common ancestor depth, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity. 
+Additionally, we suggest the adoption of four phylogeny measurements from biology: depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
 We demonstrate the use of each metric on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths, confirming our intuitions about what they can tell us about evolutionary dynamics.
@@
 
-Evolution is a collective effect of many smaller events such as replication, variation, and competition that occur on a much finer-grained temporal scale.
-While evolution's emergent nature is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results.
-In computational evolutionary systems, we can collect data to help untangle these mechanisms.
-In practice, though, gathering this data can be challenging.
-Moreover, because evolutionary effects can be witnessed on a vastly different scale from their causes, the quantity of such data can be overwhelming.
-Both of these problems can be mitigated with a standardized set of diagnostic metrics that summarize shorter-term evolutionary dynamics within a population.
-Here, we suggest a suite of such metrics that operate on lineages and phylogenies.
-We then use experimental results to provide an intuition for what they can tell us about evolution.
+Evolution is a collective effect of many smaller events such as replication, variation, and competition that occur on a fine-grained temporal scale.
+While evolution's emergent nature %is a large part of why it is so
+can be
+fascinating, it also presents challenges to studying the short-term mechanisms that, in aggregate, govern long-term results.
+In computational evolutionary systems, we can theoretically collect data to help untangle these mechanisms.
+In practice, however, %gathering this data can be challenging.
+%Moreover, because evolutionary effects can be witnessed on a vastly different scale from their causes, the quantity of such data can be overwhelming.
+the sheer number of constituent events produce an overwhelming quantity of data.
+%Both of these problems can be mitigated with
+In response, we have developed
+a standardized set of diagnostic metrics that summarize short-term evolutionary dynamics within a population.
+Here, we describe a suite of such metrics that operate on lineages and phylogenies, and provide experimental results to help develop an intuition for what they can tell us about evolution.
 
-A lineage describes a continuous line of descent, linking ancestors and their descendants.
+A lineage describes a continuous line of descent, linking %ancestors and their descendants. 
+parents and offspring in an unbroken chain from an original ancestor.
 A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
-Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in modern wet lab experiments, allowing us to replay the tape of life in exacting detail and to tease apart the evolutionary recipe for whatever complex feature or phenomenon we happen to be interested in~\citep{mcphee_using_2016}.
-In a more famous example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}.
+Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for
+%fine-grained lineage tracking at resolutions that are 
+perfect lineage tracking at a level of granularity that is
+impossible in modern wet lab experiments.
+These data allow us to replay the tape of life in exacting detail and to tease apart the evolutionary recipe for whatever complex feature or phenomenon we happen to be interested in~\citep{mcphee_using_2016}.
+In 
+%a more famous
+one notable
+example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}.
 %@SPJ: 'More famous example' sounds a bit weird to me
 
-Yet, tracking the full details of a single lineage, much less a population of lineages, can be computationally expensive and will inevitably generate an unwieldy amount of data that can be challenging to visualize and interpret~\citep{mcphee_visualizing_2016}.
+Yet, tracking the full details of a single lineage, much less a population of lineages, can be computationally expensive and will inevitably generate an unwieldy amount of data that can be challenging to visualize or interpret~\citep{mcphee_visualizing_2016}.
 Summary statistics can help alleviate these issues by enabling the user to focus on aggregate trends across a population rather than needing to examine each individual's lineage.
-The question is how to effectively summarize what is essentially a path through fitness space.
-A useful abstraction is to treat the path as a sequence of states.
-Here, we will use phenotypes and genotypes as the states in the sequence, but we could just as easily use some other descriptor of the lineage's position in the fitness landscape at a given point in time.
+The question is how to effectively summarize %what is essentially
+a path through fitness space.
+One useful abstraction is to treat the path as a sequence of states.
+Here, we use phenotypes and genotypes as the states in the sequence, but we could just as easily use some other descriptor of the lineage's position in the fitness landscape at a given point in time.
 With this abstraction in hand, a few metrics are easily formalized: the number of unique states, the number of transitions between states, and the amount of time spent in each state.
 Additionally, we may care about how the transitions between states happened.
-What mutations lead to them?
+What mutations led to them?
 Were those mutations beneficial, deleterious, or neutral at the time?
-Here, we choose a subset of these metrics that we expect will be particularly useful to explore further.
+These mutations are particularly notable because they did not simply appear briefly, but stood the test of time, leaving descendants in the final population.
+%@CAO: Does the sentence I added above make sense?
+My point being that these aren't just a random sampling of possible mutations that we are considering.
+It's almost worth an extra sentence to point out that any deleterious mutations on the lineage are of special interest since they made it into the lineage despite being actively selected against.
+% @AML: I think so. My interpretation: all mutations (beneficial, deleterious, neutral, etc., etc.) that appear on a lineage of an extant individual are particularly notable. 
+%Here, we choose a subset of these metrics that we expect will be broadly useful to explore further.
+Here, we explore a subset of these metrics that we expect will be broadly useful.
 
-Whereas a lineage represents the evolutionary history of a single individual, a phylogeny represents the evolutionary history of an entire population.
+Whereas a lineage recounts the evolutionary history of a single individual, a phylogeny details the evolutionary history of an entire population.
 Measurements that summarize phylogenies can provide useful insight into population-level evolutionary dynamics, such as diversification and co-existence among different clades.
@@
 
-Here, we present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility.
-Additionally, we suggest the adoption of four phylogeny measurements from biology: most-recent common ancestor depth, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
-For each metric, we discuss its application and our intuitions for what it can tell us about evolution.
-We evaluate our intuitions on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths.
-For this work, we restrict our attention to asexually reproducing populations; however, we make suggestions as to how these metrics can be extended to sexual populations.
+%Here, we present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility.
+Additionally, we suggest the adoption of four phylogeny measurements from biology: depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
+Here, we present three types of lineage measurements and suggest adopting four phylogeny measurements from biology; these are lineage length, mutation accumulation, phenotypic volatility, depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
+For each metric, we discuss its application and our intuition for what it can tell us about evolution.
+We evaluate our intuition on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths.
+For this work, we restrict our attention to asexually reproducing populations; however, we suggest how these metrics can extend to sexual populations.
 
@@
 %@ELD: I like the idea of putting it in that context. I was just suggesting we replace "replicator" with "selected"
-%@AML: Put the parantheticals back.
+%@AML: Put the parentheticals back.
 
@@
 
-% TODO
-
 % Define/describe
@@
 % - Neutral landscapes (expect low volatility)
-
-% Here, we...
 
@@
 Organisms selected to reproduce did so asexually.
-Each value in an offspring's genome was mutated by adding noise given by a normal distribution with a mean of 0; the `mutation rate' of a treatment defined the standard deviation used to defined this normal distribution and was given as a proportion of the test problem's domain.
-We prevented mutations from causing a value from exceeding the valid domain of the given problem. 
+Each value in an offspring's genome was mutated by adding noise given by a normal distribution with a mean of 0; the `mutation rate' of a treatment defined the standard deviation used to define this normal distribution and was given as a proportion of the test problem's domain.
+We prevented mutations from causing a value to exceed the valid domain of the given problem. 
 For each problem and tournament size, we evolved populations at eight mutation rates: $1\text{e-}08$, $1\text{e-}07$, $1\text{e-}06$, $1\text{e-}05$, $1\text{e-}04$, $1\text{e-}03$, $1\text{e-}02$, and $1\text{e-}01$.
@@
 
-\begin{figure*}
-\includegraphics[width=7in]{figs/all_ts.png}
+\begin{figure}
+\includegraphics[width=3.5in]{figs/all_ts.png}
 \caption{Values of example metrics across different tournament sizes for each of the four problems.
@@
 \label{fig:ts}
-\end{figure*}
+\end{figure}
 
@@
 However, the effect of mutation rate on the relationship between tournament size and dominant deleterious steps is complex, particularly for composition function 2 (see Supplementary Material).
-These trends likely share a common cause with the thresholding effect evident in Figure \ref{fig:mutation_rate}, where the number of deleterious steps along the dominant lineage abruptly climbs between mutation rates of 0.0000001 and .00001 and remains relatively flat over other mutation rates.
+These trends likely share a common cause with the thresholding effect evident in Figure \ref{fig:mutation_rate}, where the number of deleterious steps along the dominant lineage abruptly climbs between mutation rates of 0.0000001 and 0.00001 and remains relatively flat over other mutation rates.
 Based on an inspection of the 3-dimensional fitness landscape visualizations, we can see that this is not an effect of lineages moving from peak-to-peak; at most mutation rates, they tend to remain on a single peak.
-Thus, we can infer that this effect is the result of a drift-like phenomenon where, at sufficiently high mutation rates, all members of the population are constantly somewhat displaced from the fitness peak.
+Thus, we can infer that this effect is the result of a drift-like phenomenon where, at sufficiently high mutation rates, all members of the population are constantly somewhat displaced from their local fitness peak.
 
@@
 
+% Extend thanks to:
+%  - Devolab; thoughful discussions and whatnot
+
+% Monies:
+%  -  GRFP: Grants DGE-1424871
+%   - NSF MT Grant: DEB-1655715. 
+%   - BEACON
+
+% Other resources:
+%   - MSU HPCC
+
 \setlength{\bibsep}{2pt}

### 2018-04-03 analysis/data_analysis.Rmd
Add confidence intervals

@@
 real_value_data$selection <- sapply(as.character(real_value_data$selection),switch,"0"="Tournament","1"="Lexicase","2"="Eco-EA", "3"="MAP-Elites", "4"="Roulette", "5"="Drift")
+real_value_data$ts <- with(real_value_data, ifelse(selection=="Drift", 1, ts))
 real_value_data$selection <- as.factor(real_value_data$selection)
@@
 #real_value_data$mutation_rate <- as.factor(real_value_data$mutation_rate)
-tournament_data <- subset(real_value_data, real_value_data$selection == "Tournament" & real_value_data$elite == 0)
+tournament_data <- subset(real_value_data, (real_value_data$selection == "Tournament" | real_value_data$selection=="Drift") & real_value_data$elite == 0)
 tournament_endpoints <- filter(tournament_data, update==5000)
+tournament_endpoints$disp_pairwise_distance <- tournament_endpoints$variance_pairwise_distance/tournament_endpoints$mean_pairwise_distance
+tournament_endpoints$disp_evo_distinct <- tournament_endpoints$variance_evolutionary_distinctiveness/tournament_endpoints$mean_evolutionary_distinctiveness
 
@@
 melted_tournament_data <- melt(tournament_endpoints, id.vars = c("update", "selection", "seed", "elite", "mutation_rate", "le", "p", "problem", "g", "ts", "TOURNAMENT_SIZE", "FITNESS_INTERVAL", "RESOURCE_SELECT__OUTFLOW", "MUTATION_STD", "SELECTION_METHOD", "RESOURCE_SELECT__RES_AMOUNT", "HINT_GRID_RES", "RESOURCE_SELECT__RES_INFLOW", "LEXICASE_EPSILON", "SYSTEMATICS_INTERVAL", "RESOURCE_SELECT__FRAC", "RESOURCE_SELECT__COST", "RESOURCE_SELECT__MAX_BONUS", "POP_SNAPSHOT_INTERVAL", "POP_SIZE", "ELITE_SELECT__ELITE_CNT", "GENERATIONS"))
-end_points_vars_of_interest <- melted_tournament_data %>%  filter(variable %in% c("mrca_depth", "dominant_deleterious_steps", "mean_evolutionary_distinctiveness", "variance_evolutionary_distinctiveness", "current_phylogenetic_diversity", "mean_pairwise_distance", "variance_pairwise_distance", "phenotypic_volatility"))
+end_points_vars_of_interest <- melted_tournament_data %>%  filter(variable %in% c("mrca_depth", "dominant_deleterious_steps", "current_phylogenetic_diversity", "mean_pairwise_distance", "phenotypic_volatility", "total_magnitude"))
 
-labels <- c(current_phylogenetic_diversity = "Phylogenetic diversity", dominant_deleterious_steps = "Dominant deleterious steps", mean_evolutionary_distinctiveness="Mean evo.
+labels <- c(current_phylogenetic_diversity = "Phylogenetic diversity", dominant_deleterious_steps = "Deleterious steps", mean_evolutionary_distinctiveness="Mean evo.
 distinctiveness", mean_pairwise_distance="Mean pairwise distance", mrca_depth="MRCA depth", variance_evolutionary_distinctiveness="Variance evo.
-distinctiveness", variance_pairwise_distance = "Variance pairwise distance", phenotypic_volatility="Phenotypic volatility")
+distinctiveness", variance_pairwise_distance = "Variance pairwise distance", phenotypic_volatility="Phenotypic volatility", total_magnitude="Mutation accumulation")
 
@@
 ```{r}
-ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$mutation_rate==.001), aes(x=ts, y=value, color=problem, group=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="line", fun.y=median)+ theme_classic() + scale_x_log10("Tournament size", breaks=c(2,4,8,16)) + scale_y_continuous("", labels=comma) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels)) + scale_color_discrete("Problem") + theme(legend.position = "bottom")
-ggsave("../figs/all_ts.png", width=7, height=5) 
+ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$mutation_rate==.001), aes(x=ts, y=value, fill=problem, color=problem, group=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Tournament size", breaks=c(1,2,4,8,16)) + scale_y_log10("", labels=scientific, breaks = scales::trans_breaks("log10", function(x) 10^x)) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels),ncol=2) + scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+ggsave("../figs/all_ts.png", width=3.5, height=4.5) 
 ```
@@
 ```{r}
-ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$ts==4), aes(x=mutation_rate, y=value, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="line", fun.y=median)+ theme_classic() + scale_x_log10("Mutation rate", labels=scientific) + scale_y_continuous("", labels=comma) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels)) +scale_color_discrete("Problem") +theme(legend.position = "bottom")
-ggsave("../figs/all_mutation_rate.png", width=7, height=5)
+ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$ts==4), aes(x=mutation_rate, y=value, fill=problem, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Mutation rate", labels=scientific) + scale_y_log10("", breaks = scales::trans_breaks("log10", function(x) 10^x), labels=scientific) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels), ncol=2) +scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+ggsave("../figs/all_mutation_rate.png", width=3.5, height=4.5)
 

### 2018-04-03 paper.tex

@@
 %@ELD: I am in favor.
-In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we intend for this work to begin a conversation within the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories.
-There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are incredibly valuable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history.
+In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we intend for this work to begin a conversation within the %computational evolution
+artificial life
+community about how we quantify, interpret, and compare observed evolutionary histories.
+There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are %incredibly valuable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history.
 However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-related metrics for computation evolution.
@@
 % Changing mutation: here's what happens when we change mutation on each problem
-\begin{figure*}
-\includegraphics[width=7in]{figs/all_mutation_rate.png}
+\begin{figure}
+\includegraphics[width=3.5in]{figs/all_mutation_rate.png}
 \caption{Values of example metrics across different mutation rates for each of the four problems.
@@
 \label{fig:mutation_rate}
-\end{figure*}
+\end{figure}
 

### 2018-04-03 paper.tex

@@
 community about how we quantify, interpret, and compare observed evolutionary histories.
-There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are %incredibly valuable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history.
+There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are %incredibly valuable
+indespensible 
+for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history.
 However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-related metrics for computation evolution.

### 2018-04-03 paper.tex

--- 2018-04-03 paper.tex
@@
 %Here, we choose a subset of these metrics that we expect will be broadly useful to explore further.
+%@ELD: I agree. It's also nice to emphasize in light of deleterious mutations being the most interesting part of our results.
 Here, we explore a subset of these metrics that we expect will be broadly useful.
@@
 community about how we quantify, interpret, and compare observed evolutionary histories.
-There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are %incredibly valuable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history.
-However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-related metrics for computation evolution.
-Indeed, ....
+There have been extensive efforts to improve our ability to visualize both lineages and phylogenies~\citep{standish_visualising_2002,burlacu_visualization_2013, mcphee_using_2016,mcphee_visualizing_2016, lalejini_evolutionary_2016}, which are %incredibly valuable
+indispensable 
+for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history.
+However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-based metrics for computational evolution.
+%Indeed, ....
 % people have done stuff with lineages
+% @CAO: For the moment I commented the "Indeed" out since I'm not sure where it's going; think the current version is a fine place to stop, though I also agree that some of the points below would be good to include if we can.
 % - often a component of a larger experiment-specific analysis or limited to qualitative descriptions/visualizations
@@
 
-Code for all of our metrics is open source and available in the Empirical library (https://github.com/devosoft/Empirical).
+Code for all of our metrics is open source and available in the Empirical library (\url{https://github.com/devosoft/Empirical}).
 Empirical is a C++ library built to facilitate writing efficient and easily sharable scientific software.
-To facilitate including useful features without excessive overhead, Empirical is header-only, meaning these metrics can be imported into an existing project with minimal effort.
+To simplify including useful features with minimal overhead, Empirical is header-only, meaning these metrics can be imported into an existing project with limited effort.
 % [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager? 
@@
 \subsection{Lineage Metrics}
-% Outlining some points to (maybe) make here; probably too bogged down in details; not even sure we want to talk about lineages as graphs:
-% - Apply metrics in context of asexual populations
-% - Establish a framework from which to think about metrics (not novel, but rarely formalized)
-% 	 - We treat lineages as sequences of states where each state represents an individual and transitions between states represent parent-offspring relationships.
-%      - More formally, lineages are directed acyclic graphs where internal nodes have in-degree = 1 (parent), out-degree = 1 (offspring).
-Original ancestor (in digital evolution, often seed of population) w/out-degree = 1.
-Extant individual w/in-degree = 1.
-%      - Transitions specify relevant variation introduced into offspring (e.g. mutations)
-%      - States specify relevant genotype and phenotype information (e.g. genome, location in space, behavior, etc) -- anything we might want to measure over a lineage
-%      - Becomes more complicated with sexual populations, but we'll stick with asexual assumption for this. 
-%  - Already, framework affords certain questions:
-%     - How many individuals are along the lineage? 
-%     - Total # and magnitude of mutations along lineage?
-%     - Distribution of mutation types (ins vs. sub/beneficial vs. deleterious)?
-%  - Further abstractions:
-%      - Sequence of genotypes along lineage (where each genotype may represent multiple individuals)
-%      - Sequence of phenotypes along lineage (where each phenotype may represent multiple individuals)
-%      - Still further abstraction: sequence of a particular aspect of genotype (e.g. genetic marker)/phenotype (e.g. location) along lineage
-%  - We discuss three classes of lineage metrics within this framework: lineage length, mutation accumulation, and phenotypic volatility. 
-%  - For each, we discuss our intuition for what they can tell us (alone and in combination) about evolution. 
-Each of the three lineage metrics we discuss --- lineage length, mutation accumulation, and phenotypic volatility --- reduces a lineage to a linear sequence of states where each state represents an individual or a continuous sequence of individuals that share some common genotypic or phenotypic characteristic of interest.
+Each of the three lineage metrics that we discuss --- lineage length, mutation accumulation, and phenotypic volatility --- reduces a lineage to a linear sequence of states where each state represents an individual or sequence of individuals that share a common genotypic or phenotypic characteristic of interest; Figure \ref{fig:lineage_metrics_cartoon} is given as a toy example to help guide our discussion of these metrics.
 While we limit our focus to three lineage metrics, this abstraction places lineages in a form suitable for a wide range of measurements, including the direct application of many data mining techniques designed to operate over sequences such as sequential pattern mining, trend analysis, \textit{et cetera} \citep{han2011data}.
 
-Only lineages consisting of asexually reproducing ancestors where genetic material is exclusively vertically transmitted can be directly abstracted as a \textit{linear sequence} of states. 
+Only asexual lineages %consisting of asexually reproducing ancestors 
+where genetic material is exclusively vertically transmitted can be directly abstracted as a \textit{linear sequence} of states.
 Sexual reproduction (and any form of horizontal gene transfer) complicates matters significantly as such lineages are more appropriately represented by trees rooted at the extant organism, branching for each contributor of genetic material.
-For each metric we present here, we limit our discussion to asexual populations; however, we suggest two approaches for generalizing these lineage metrics to sexual populations: apply a lossy compression to reduce sexual lineages into linear sequences or extend our application of the metrics to operate over non-linear (tree) structured lineages.
-Sexual lineages can be compressed into linear sequences of states by modeling sexual reproduction events as asexual reproduction events, designating one parent to be a part of the lineage and considering the genetic contributions of other parents as sources of genetic variation (mutations).
+% For the metrics we present here, we limit our discussion to asexual populations; however, we suggest two approaches for generalizing these lineage metrics to sexual populations: apply a lossy compression to reduce sexual lineages into linear sequences or extend our application of the metrics to operate over non-linear (tree) structured lineages.
+One possibility is to compress sexual lineages into linear sequences of states by modeling sexual reproduction events as asexual reproduction events, designating one parent to be a part of the lineage and considering the genetic contributions of other parents as sources of genetic variation (mutations).
 The primary downside to this approach is its lossy-ness (i.e.
@@
 One such approach would be to consider all possible ancestor paths for an extant individual, calculating a given metric for each of them and then averaging the resulting values together.
+Another approach would be to divide an organism into its constituent parts that are inherited atomically (such as genes or instructions, depending on the representation); an organism would then be viewed as a collection of lineages rather than a single one.
 Assessing the efficacy of these and potentially other approaches would be a useful line of research to pursue in the future.
 
+\begin{figure*}[t]
+\includegraphics[width=\textwidth]{figs/lineage_metrics_cartoon.png}
+\caption{\small Panels (A), (B), (C), and (D) each show different ways of representing a lineage.
+The lineage has accumulated three mutations: one reverse mutation and two substitutions.
+The phenotype of ancestors changed three times.
+In (A), each ancestral state along the lineage represents a single individual.
+Here, the lineage length is equal to the number of generations spanned by the lineage (eight).
+In (B), ancestral states represent the sequence of genotypes along the lineage.
+Here, the lineage length is equal to four.
+In (C) ancestral states represent the sequence of phenotypes along the lineage.
+Here, the lineage length is equal to the number of times a different phenotype is expressed along the lineage (three).
+In (D), ancestral states are given by a particular phenotypic characteristic.
+Here, lineage length is equal to two.}
+\label{fig:lineage_metrics_cartoon}
+\end{figure*}
+
 \subsubsection{Lineage Length}
-% Define/explanation
-% 	- Within context of framework, lineage length is simply measure of number of states in the graph/sequence.
-%   - What this means will depend on what states represent?
-%       - number of individuals along lineage 
-%             - More interesting in experiments where generations are not dictated by system, but rather by life history strategies of organisms
-%             - Give you a diagnostic for replication rate (r vs. k strategy)
-%      - number of genotypes along lineage
-%            - length = 1: no mutations
-%            - length = number of individuals: mutation in every replication
-%            - Intuitively, should be higher at high mutation rates, lower at low mutation rates. 
-%      - number of phenotypes along lineage
-%            - low: evidence for static environment
-%            - high: might be indicative of a changing environment or ecological interactions
-%            - if number of genotypes is high, but number of phenotypes low, probably in a neutral fitness landscape
-%      - various subsets of genotype/phenotype
-%           - location in space (environment type): useful for looking at lineage through heterogeneous landscape
-%           - change in particular phenotypic trait
-%           - change in genetic markers
-%           - Distributions of above two lengths can tell you useful things about drift, selection, etc.
-
 Lineage length describes the number of ancestral states traversed by a lineage.
 If a state represents a single individual along a lineage, lineage length is a measure of the number of generations on a lineage.
-This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined but instead determined by the life history strategies of organisms.
+This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined, but instead determined by the life history strategies of organisms.
 In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates (\textit{e.g.} r-selected lineage) and shorter lineages imply slower replication rates (\textit{e.g.} K-selected lineage).
@@
 Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.
+%@CAO There are various other things it could tell us, such as additional co-evolutionary pressures that are not reflected in the raw calculation of fitness.
+Interestingly, using these metrics in combinaton with local fitness landscape analyses would do the trick.
+But that's for another paper.
+Indeed, I think it would be nice to put out a series of papers of useful analysis techniques and then do a master paper that reviews them all and demonstrates them in the context of an Avida-like system AND a very different digital evolution system (Markov Brains?)
 
@@
 
-Phenotypic volatility addresses the rate at which the lineage changes phenotype (although the same concept can be applied to other types of state).
+Phenotypic volatility addresses the rate at which phenotype changes as you move down a lineage (although the same concept can be applied to specific phenotypic traits or other types of state).
 In systems with discrete/categorical phenotypes, this can be measured by summing the number of times the phenotype changes.
 A related but subtly different measurement in such systems is the number of unique phenotypes on a lineage.
-In most cases, these values will be very similar.
-A discrepancy between them, however, would be a clear signal of interesting dynamics.
-Specifically, it would suggest that the lineage was in a portion of the landscape where it could rapidly switch among different phenotypes.
-Such behavior could be indicative of some form of evolutionary bet-hedging~\citep{beaumont2009experimental}.
+In most cases, these values will be similar; a discrepancy
+%between them, however, would be a clear signal of interesting dynamics. Specifically, it 
+would suggest that the lineage was %in a portion of the landscape where it could rapidly switch among different 
+cycling through a set of
+phenotypes.
+Such behavior could, for example, be indicative of some form of evolutionary bet-hedging~\citep{beaumont2009experimental}.
 % @AML: There are a whole lot of citations for bet-hedging..
@@
 In some cases, it may be desirable to smooth out the noise inherent in a real-valued phenotype.
-We can do so by instead taking the variance of the rolling mean of fitness, to more closely approximate the idea of measuring phase transitions.
+We can do so by instead taking the variance of the moving average of fitness, to more closely approximate the idea of measuring phase transitions.
 % SPJ: What is meant by 'rolling mean' ?
 %@ELD: Would "moving average" be more familiar? 
+%@ELDL Also there turns out not to be much of a difference, so we could just not mention that approach.
+% @SPJ: I don't think I have heard either of those terms. If we need space, it may be something to cut?
+% @CAO: I suspect "moving average" is the more popular term, but I suspect people would figure out what we meant by either.
+% @ELD: Okay, I changed it, but also this sentence is the first to go if we run into space issues.
 
@@
 Instead, we need to come up with ways to generate useful summaries.
-There are two main approaches to doing so: 1) choose a small number of representative lineages from a given time point, 2) collect summary statistics about the distribution of metric values across the population.
+There are two main approaches to doing so: 1) choose a small number of representative lineages from a given time point, or 2) collect summary statistics about the distribution of metric values across the population.
 
-A single representative lineage can be chosen by selecting the lineage of the dominant organism (either the most fit or the more numerous; here we use the most fit).
-In populations with a diversity of coexisting strategies, this approach can be uninformative as the dominant lineage is not necessarily representative of all successful lineages.
-An alternative in these scenarios is to filter out lineages that do not have offspring some predetermined number of generations later as such lineages were likely not representative of an important subset of the population.
+A single lineage can be chosen by selecting the lineage of %the dominant
+a representative
+organism (either the most fit or the more numerous; here we use the most fit).
+In populations where diverse strategies coexist, this approach can be uninformative as any one lineage is unlikely to be representative of all successful lineages.
+One alternative %in these scenarios
+is to filter out lineages that do not have offspring some predetermined number of generations later as such lineages were likely not representative of an important subset of the population.
 Still, any approach based on measuring only a subset of lineages can be challenging to interpret when the current dominant lineage (or lineages) is replaced with a different one; such changes can introduce a discontinuity if the value is being measured over time.
@@
 
-In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be very similar to the average lineage in the population, as most of the population will be closely related.
+In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be similar to the average lineage,
+%in the population,
+as most of the population will be closely related.
 When the population contains more phylogenetic diversity, however, it is possible for the dominant lineage to differ from the mean.
-The nature of such differences is likely informative about the evolutionary dynamics occurring in the population.
+Of course, the nature of such differences is likely informative about the evolutionary dynamics occurring in the population.
 %Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. 
@@
 These metrics operate on entire phylogenies rather than single lineages within a population.
-Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics.
-Because they factor in the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage.
-On the other hand, because most lineages in the population share a substantial amount of history, phylogeny metrics can often be calculated more rapidly than lineage metrics can be calculated across an entire population.
+%Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics.
+Thus, there is no need to identify a representative organism or lineage.
+Because they %factor in 
+use date from
+the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage.
+On the other hand, because most lineages in the population share substantial history, phylogeny metrics can usually be calculated more rapidly than lineage metrics %can be calculated across
+for an entire population.
 Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc.
 Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}.
-% @SPJ: could reword to 'Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}.',  might be more clear?
-%@ELD: sounds good!
 
-A standard technique for saving memory and time when working with lineages in computational systems is to ``prune" them, removing any taxa that are not either currently alive or the ancestor of taxa that are alive.
+A standard technique for saving memory and time when working with phylogenties in computational systems is to ``prune" them, removing 
+%any taxa that are not either currently alive or the ancestor of taxa that are alive.
+% either dead (extinct) branches, or ancestors common to the entire population.
+%@CAO: (NEVERMIND - SIMPLIFYING FURTHER... Does the substition I made above make sense?  I'm not quite sure what was meant by "the ancestor of taxa that are alive".
+dead (extinct) branches.
 Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies.
-Thus, for the remainder of this paper, we will assume we are only working with pruned phylogenies.
+Thus, for the remainder of this paper, we will assume we are working with pruned phylogenies.
 
@@
 
-An important distinction between phylogenies in natural versus computational systems to be aware of is that in natural systems phylogenies are generally inferred from extant taxa, whereas in computational systems they are directly measured.
-The inferred phylogenies used in natural systems do not contain internal nodes unless those nodes are branch points.
+An important distinction between phylogenies in natural versus computational systems to be aware of is that natural phylogenies are generally inferred from extant taxa, whereas computational phylogenies are directly recorded.
+The inferred phylogenies used in natural systems do not contain internal nodes except at branch points.
 They also do not contain history prior to the most recent common ancestor (MRCA) of all organisms in the extant population.
@@
 
-The depth of the MRCA (i.e.
-how many steps from the beginning of time did it occur?) is a useful metric because of the relatively high ratio of informativeness to ease of calculation.
-A more recent MRCA implies more frequent selective sweeps and less long-term stable coexistence between clades.
-The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require keeping track of it already.
+The depth of the MRCA %(i.e. how many steps from the beginning of time did it occur?) 
+(i.e., the number of steps it is from the original ancestor)
+%is a useful metric because of the relatively high ratio of informativeness to ease of calculation.
+is an informative metric and is easy to calculate.
+A recent MRCA implies frequent selective sweeps and less long-term stable coexistence between clades.
+%The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require keeping track of it already.
 Measuring the frequency with which the MRCA changes (i.e.
@@
 This scenario is particularly likely when the population size is changing over time.
-A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never see a change in the MRCA after the very beginning of evolution (which is informative as a way of detecting stably coexisting branches in the phylogeny).
-% @SPJ: change to 'The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require ...' (i.e remove 'so' and 'most any')
-% @SPJ: The above sentence might need use some support (why is it informative?)
-%@ELD: Better?
+A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never change its MRCA after the very beginning of evolution %(which is informative as a way of detecting stably coexisting branches in the phylogeny).
+(which at least allows us to detect stably coexistance in the population).
 
@@
 
-Measurements of phylogenetic richness quantify the total amount of evolutionary history that is contained in a set of taxa.
-The most traditional metric of phylogenetic richness is the one simply titled ``Phylogenetic Diversity", which is calculated as the number of the nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}.
+Measurements of phylogenetic richness quantify the total amount of evolutionary history contained in a set of taxa.
+The most traditional metric of phylogenetic richness is the one simply titled ``Phylogenetic Diversity", which is calculated as the number of nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}.
 Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}.
-A third approach is to calculate evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, and sum it across all extant taxa~\citep{tucker_guide_2017}.
+A third approach is to sum evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, across all extant taxa~\citep{tucker_guide_2017}.
 
@@
 
-Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other and are often the means of values that can be calculated for individual taxa.
-For instance, one option is to calculate the mean of the pairwise distances between all taxa in the population~\citep{webb_exploring_2000}.
-Similarly, phylogenetic divergence can also be calculated by calculating the evolutionary distinctiveness of each taxon in the population (as alluded to in the previous section) and taking the average of these values.
+Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other and are often %the means of values that can be calculated for
+averaged across
+individual taxa. For example, one option is to 
+%calculate the mean of the pairwise distances between 
+average the pairwise distances across
+all taxa in the population~\citep{webb_exploring_2000}. Similarly, phylogenetic divergence can be calculated by 
+%calculating the evolutionary distinctiveness of each taxon in the population (as alluded to in the previous section) and taking the average of these values.
+% @CAO: "can be calculated by calulating" ?
+averaging the evolutionary distinctiveness across each taxon in the population.
+%(as alluded to in the previous section).
 
@@
 \includegraphics[width=3.5in]{figs/landscapes.png}
-\caption{The fitness landscapes used in this experiment: A) Himmelblau, B) Six-humped Camel Back, C) Shubert, and D) Composition Function 2.
+\caption{\small The fitness landscapes used in this experiment: A) Himmelblau, B) Six-humped Camel Back, C) Shubert, and D) Composition Function 2.
 Interactive versions available at \url{https://emilydolson.github.io/fitness_landscape_visualizations}.}
@@
 We chose a diverse subset of these functions (Himmelblau, Shubert, Composition Function 2, and Six-Humped Camel Back) as our test problems in order to gain a broad understanding of our metrics.
-We used the implementations of these problems at https://github.com/mikeagn/CEC2013/ (C++ for fitness calculations during evolution, Python for post-hoc analysis).
+We used the implementations of these problems at \url{https://github.com/mikeagn/CEC2013} (C++ for fitness calculations during evolution, Python for post-hoc analysis).
 Figure \ref{fig:landscapes} shows the fitness landscapes defined by each of our four chosen test problems.
@@
 For all experimental conditions, we ran ten replicates, each with a unique random number seed. 
-Our experiment is implemented using the Empirical library; our implementation can be found at https://github.com/stevenjson/ALife2018-Lineage.
+Our experiment is implemented using the Empirical library; our implementation can be found at \url{https://github.com/stevenjson/ALife2018-Lineage}.
 % @AML: We might want to rename the repository to match whatever the title of this paper ends up being (e.g.
@@
 \subsection{3D visualizations}
+
+\begin{figure}
+\includegraphics[width=3.5in]{figs/shubert_landscape.png}
+\caption{\small A close-up on two adjacent peaks in the Shubert function fitness landscape.
+Lineages are depicted as paths fading from white to black over evolutionary time.
+The lienages shown here evolved under a mutation rate of 0.01.
+A) Was evolved using a tournament size of 2, whereas B) was evolved using a tournament size of 16.
+These figures neatly illustrate how increased tournament size keeps the lineage near the tops of the peaks.}
+\label{fig:lineageviz}
+\end{figure}
+
 % Trim dramatically
@@
 In order to make these metrics useful in the long run, it is critical that we have an accurate understanding of how various measurements correspond to the actual behavior of lineages.
-The most direct way to confirm our expectations is to visualize the path that each lineage takes across the fitness landscape, mapping the x, y, and z (fitness) coordinates of each ancestor of each member of the population.
+The most direct way to confirm our expectations is to visualize the path that each lineage takes across the fitness landscape, mapping the x, y, and z (fitness) coordinates of each ancestor of each member of the population~\citep{virgo_lineage_2017}.
 Creating such a visualization entails incorporating a large quantity of information into a limited space.
 When projected onto two dimensions, lineages can obscure parts of the fitness landscape (and each other).
-To mitigate this problem, we built a thee-dimensional visualization using the A-Frame Framework [cite].
+To mitigate this problem, we built a thee-dimensional visualization (see Figure \ref{fig:lineageviz}) using the A-Frame Framework, available at \url{https://www.aframe.io}.
 
@@
 For the data interpretation in this paper, we used an Oculus Rift to provide us with fine-grained control of which part of the visualization we were looking at.
+Our full visualization, complete with data, can be viewed on the web or in VR at \url{https://stevenjson.github.io/ALife2018-Lineage/analysis/fitness_landscape_visualizations/}.
 
@@
 Specifically, we used the ggplot2 library for all graphs included in this paper~\citep{wickham_ggplot2:_2009}.
-All analysis scripts are available at https://github.com/stevenjson/ALife2018-Lineage.
+All analysis scripts are available at \url{https://github.com/stevenjson/ALife2018-Lineage}.
 
@@
 \includegraphics[width=3.5in]{figs/all_mutation_rate.png}
-\caption{Values of example metrics across different mutation rates for each of the four problems.
+\caption{\small Values of example metrics across different mutation rates for each of the four problems.
+All lineage-based metrics are calculated on the lineage of the fittest organism at the final time point (the dominant); population-level means behaved similarly.
 All experiments shown here used a tournament size of 4.
-Circles are medians and vertical lines show inter-quartile range.}
+Circles are medians, vertical lines show inter-quartile range, and shaded area is a bootstrapped 95\% confidence interval around the mean.
+Note that both axes are on log scales.}
 \label{fig:mutation_rate}
@@
 \includegraphics[width=3.5in]{figs/all_ts.png}
-\caption{Values of example metrics across different tournament sizes for each of the four problems.
-All experiments shown here used a mutation rate of .001.
-Circles are medians and vertical lines show inter-quartile range.}
+\caption{\small Values of example metrics across different tournament sizes for each of the four problems.
+All experiments shown here used a mutation rate of 0.001.
+All lineage-based metrics are calculated on the lineage of the fittest organism at the final time point (the dominant); population-level means behaved similarly.
+Circles are medians, vertical lines show inter-quartile range, and shaded area is a bootstrapped 95\% confidence interval around the mean.
+Note that both axes are on log scales.}
 \label{fig:ts}
@@
 
-Surprisingly, there is no clear effect of tournament size on the count of deleterious steps along the dominant lineage.
+Surprisingly, there is no clear effect of tournament size on the count of deleterious steps along the dominant lineage (as evidenced by the fact that the confidence intervals are all overlapping).
 Values for all selection schemes and tournament sizes hover near 2,500, meaning that a deleterious step is taken in roughly half of the 5000 generations.
@@
 
-The primary goal of this work is two-fold: 1) to communicate a set of useful metrics aimed at improving our capacity to quantitatively understand evolutionary histories in digital evolution experiments, and 2) to spark a conversation in the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories.
+The primary goal of this work is two-fold: 1) to suggest a set of metrics that will improve our capacity to quantitatively understand evolutionary histories in digital evolution experiments, and 2) to spark a conversation in the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories.
 With feedback from the community, we plan to expand the presented suite of lineage and phylogeny metrics, compiling accessible descriptions and examples of each metric.
 
-In this work we demonstrated the suite of lineage and phylogeny metrics on a set of toy problems where evolving digital organisms are overly simplistic.
-Next, we plan to apply these metrics to more complex digital organisms that we evolve in a variety of qualitatively different environments where we would expect to observe a wide range of evolutionary dynamics.
+We have demonstrated that the suite of lineage and phylogeny metrics behaves reasonably on a set of toy problems with simplistic organisms.
+Having established baseline expectations for their responses to common conditions, our next step is to apply these metrics to more complex scenarios: populations of digital organisms that we evolve in a variety of qualitatively different environments where we would expect to observe a wide range of evolutionary dynamics.
+It is under these conditions that we expect that true value of these metrics to become clear.
 
@@
 %  - Devolab; thoughful discussions and whatnot
+%@ELD: Anyone specific?
 
@@
 %  -  GRFP: Grants DGE-1424871
-%   - NSF MT Grant: DEB-1655715. 
+%   - NSF MT Grant: DEB-1655715. (for funds to develop Empirical tools)
 %   - BEACON
@@
 %   - MSU HPCC
+
+We thank all members of the MSU Digital Evolution Lab for their helpful comments and suggestions on early drafts of this manuscript.
+This research has been supported by the National Science Foundation (NSF) BEACON Center under Cooperative Agreement DBI-0939454, by NSF Graduate Research Fellowships to ED and AL under Grant No.
+DGE-1424871, by NSF Grant No.
+DEB-1655715, which provided support for development of the software tools, and by Michigan State University through computational resources provided by the Institute for Cyber-Enabled Research and the Digital Scholarship Lab.
+Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the NSF.
+% @CAO: If needed, we can collapse a lot of the NSF grants above into a litany of grant numbers making the paragraph much shorter.
 
@@
 \end{document}
+
+% ===== Where old comments go to die:  ======
+
+% Outlining some points to (maybe) make here; probably too bogged down in details; not even sure we want to talk about lineages as graphs:
+% - Apply metrics in context of asexual populations
+% - Establish a framework from which to think about metrics (not novel, but rarely formalized)
+% 	 - We treat lineages as sequences of states where each state represents an individual and transitions between states represent parent-offspring relationships.
+%      - More formally, lineages are directed acyclic graphs where internal nodes have in-degree = 1 (parent), out-degree = 1 (offspring).
+Original ancestor (in digital evolution, often seed of population) w/out-degree = 1.
+Extant individual w/in-degree = 1.
+%      - Transitions specify relevant variation introduced into offspring (e.g. mutations)
+%      - States specify relevant genotype and phenotype information (e.g. genome, location in space, behavior, etc) -- anything we might want to measure over a lineage
+%      - Becomes more complicated with sexual populations, but we'll stick with asexual assumption for this. 
+%  - Already, framework affords certain questions:
+%     - How many individuals are along the lineage? 
+%     - Total # and magnitude of mutations along lineage?
+%     - Distribution of mutation types (ins vs. sub/beneficial vs. deleterious)?
+%  - Further abstractions:
+%      - Sequence of genotypes along lineage (where each genotype may represent multiple individuals)
+%      - Sequence of phenotypes along lineage (where each phenotype may represent multiple individuals)
+%      - Still further abstraction: sequence of a particular aspect of genotype (e.g. genetic marker)/phenotype (e.g. location) along lineage
+%  - We discuss three classes of lineage metrics within this framework: lineage length, mutation accumulation, and phenotypic volatility. 
+%  - For each, we discuss our intuition for what they can tell us (alone and in combination) about evolution.
+
+% Define/explanation
+% 	- Within context of framework, lineage length is simply measure of number of states in the graph/sequence.
+%   - What this means will depend on what states represent?
+%       - number of individuals along lineage 
+%             - More interesting in experiments where generations are not dictated by system, but rather by life history strategies of organisms
+%             - Give you a diagnostic for replication rate (r vs. k strategy)
+%      - number of genotypes along lineage
+%            - length = 1: no mutations
+%            - length = number of individuals: mutation in every replication
+%            - Intuitively, should be higher at high mutation rates, lower at low mutation rates. 
+%      - number of phenotypes along lineage
+%            - low: evidence for static environment
+%            - high: might be indicative of a changing environment or ecological interactions
+%            - if number of genotypes is high, but number of phenotypes low, probably in a neutral fitness landscape
+%      - various subsets of genotype/phenotype
+%           - location in space (environment type): useful for looking at lineage through heterogeneous landscape
+%           - change in particular phenotypic trait
+%           - change in genetic markers
+%           - Distributions of above two lengths can tell you useful things about drift, selection, etc.

### 2018-04-03 paper.tex
Update on Overleaf.

@@
 
-% *****************
-%  TODOs:
-% *****************
-% [ ] Writing: Abstract
-% [ ] Writing: Results & discussion
-% [ ] Writing: Conclusions
-% [ ] Big 'ol citation pass: need to work appropriate citations into paper
+\newcommand{\reduceCaptionPadding}{\setlength{\belowcaptionskip}{-10pt}}
+
+\renewcommand\floatpagefraction{.9}
+\renewcommand\dblfloatpagefraction{.9} % for two column documents
+\renewcommand\topfraction{.9}
+\renewcommand\dbltopfraction{.9} % for two column documents
+\renewcommand\bottomfraction{.9}
+\renewcommand\textfraction{.1}
 
@@
 %  - A Suite of Metrics for Understanding Evolutionary Dynamics/Histories
-\title{Understanding fine temporal scale evolutionary dynamics with lineage metrics}
+%  - Quantifying Evolutionary Histories 
+%  - A Suite of Metrics for Quantifying Evolutionary Histories
+\title{Quantifying the tape of life: Ancestry-based metrics provide insights and intuition about evolutionary dynamics}
 \author{Emily Dolson$^{1,2,3}$, Alexander Lalejini$^{1,2, 3}$, Steven Jorgensen$^{1,2}$ \and Charles Ofria$^{1,2, 3}$ \\
@@
 \begin{abstract}
-Fine-scale evolutionary dynamics can be challenging to tease out when focused on broad brush stokes of whole populations over long timespans.
+Fine-scale evolutionary dynamics can be challenging to tease out when focused on broad brush strokes of whole populations over long timespans.
 We propose a suite of diagnostic metrics that operate on lineages and phylogenies in digital evolution experiments with the aim of improving our capacity to quantitatively explore the nuances of evolutionary histories in digital evolution experiments.
@@
 Evolution is a collective effect of many smaller events such as replication, variation, and competition that occur on a fine-grained temporal scale.
-While evolution's emergent nature %is a large part of why it is so
-can be
-fascinating, it also presents challenges to studying the short-term mechanisms that, in aggregate, govern long-term results.
+While evolution's emergent nature can be fascinating, it also presents challenges to studying the short-term mechanisms that, in aggregate, govern long-term results.
 In computational evolutionary systems, we can theoretically collect data to help untangle these mechanisms.
-In practice, however, %gathering this data can be challenging.
-%Moreover, because evolutionary effects can be witnessed on a vastly different scale from their causes, the quantity of such data can be overwhelming.
-the sheer number of constituent events produce an overwhelming quantity of data.
-%Both of these problems can be mitigated with
-In response, we have developed
-a standardized set of diagnostic metrics that summarize short-term evolutionary dynamics within a population.
-Here, we describe a suite of such metrics that operate on lineages and phylogenies, and provide experimental results to help develop an intuition for what they can tell us about evolution.
+In practice, however, the sheer number of constituent events produce an overwhelming quantity of data.
+In response, we have developed a standardized suite of diagnostic metrics to summarize short-term evolutionary dynamics within a population by measuring lineages and phylogenies.
+Here, we describe these metrics and provide experimental results to develop an intuition for what they can tell us about evolution.
 
-A lineage describes a continuous line of descent, linking %ancestors and their descendants. 
-parents and offspring in an unbroken chain from an original ancestor.
+A lineage describes a continuous line of descent, linking parents and offspring in an unbroken chain from an original ancestor.
 A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
-Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for
-%fine-grained lineage tracking at resolutions that are 
-perfect lineage tracking at a level of granularity that is
-impossible in modern wet lab experiments.
-These data allow us to replay the tape of life in exacting detail and to tease apart the evolutionary recipe for whatever complex feature or phenomenon we happen to be interested in~\citep{mcphee_using_2016}.
-In 
-%a more famous
-one notable
-example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}.
-%@SPJ: 'More famous example' sounds a bit weird to me
+Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for perfect lineage tracking at a level of granularity that is impossible in modern wet lab experiments.
+These data allow us to replay the tape of life in precise detail and to tease apart the evolutionary recipe for any phenomenon we are interested in~\citep{mcphee_using_2016}.
+In one notable example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}.
 
@@
 Summary statistics can help alleviate these issues by enabling the user to focus on aggregate trends across a population rather than needing to examine each individual's lineage.
-The question is how to effectively summarize %what is essentially
-a path through fitness space.
+The question is how to effectively summarize a path through fitness space.
 One useful abstraction is to treat the path as a sequence of states.
@@
 Thus, they are useful indicators of the presence of various types of eco-evolutionary dynamics.
-%@AML: I think the next bit (just moved it) might belong in the metrics section. Instead, we might want a little more elaboration on each of the categories of metrics.
-%@ELD: Is that too handwavy?
-%@AML: Nah! Handwaving is just what we needed!
 
-%Here, we present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility.
-Additionally, we suggest the adoption of four phylogeny measurements from biology: depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
 Here, we present three types of lineage measurements and suggest adopting four phylogeny measurements from biology; these are lineage length, mutation accumulation, phenotypic volatility, depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
@@
 
-%@AML:  It might be nice to wrap this section up with a little bit of a 'call to action'/'why should you care about this as a community'
-%@ELD: I am in favor.
-In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we intend for this work to begin a conversation within the %computational evolution
-artificial life
-community about how we quantify, interpret, and compare observed evolutionary histories.
-There have been extensive efforts to improve our ability to visualize both lineages and phylogenies~\citep{standish_visualising_2002,burlacu_visualization_2013, mcphee_using_2016,mcphee_visualizing_2016, lalejini_evolutionary_2016}, which are %incredibly valuable
-indispensable 
-for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history.
+In addition to demonstrating a range of metrics that are %immediately applicable and
+useful to digital evolution research, we intend for this work to begin a conversation within the %computational evolution
+artificial life community about how we quantify, interpret, and compare observed evolutionary histories.
+There have been extensive efforts to improve our ability to represent and visualize both lineages and phylogenies~\citep{standish_visualising_2002,burlacu_visualization_2013, mcphee_using_2016,mcphee_visualizing_2016, lalejini_evolutionary_2016}, which are indispensable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history.
 However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-based metrics for computational evolution.
@@
 Empirical is a C++ library built to facilitate writing efficient and easily sharable scientific software.
-To simplify including useful features with minimal overhead, Empirical is header-only, meaning these metrics can be imported into an existing project with limited effort.
+%To simplify including useful features with minimal overhead,
+Empirical is a header-only library, so adding these metrics to an existing project has minimal overhead.
 % [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager? 
@@
 \subsection{Lineage Metrics}
+
+\begin{figure*}[t]
+\includegraphics[width=\textwidth]{figs/lineage_metrics_cartoon.png}
+\reduceCaptionPadding
+\caption{\small Four methods of representing a lineage.
+This example lineage has accumulated three mutations (one reverse mutation and two substitutions) and gone through three distinct phenotypes.
+In (A), each state along the lineage represents a single individual; lineage length is the number of generations spanned by the lineage (eight).
+In (B), states represent the sequence of genotypes along the lineage, reducing lineage length to four.
+In (C) states represent the sequence of phenotypes along the lineage; lineage length is the number of times a different phenotype is expressed (three).
+In (D), states are a particular phenotypic characteristic; here, lineage length is two.}
+\label{fig:lineage_metrics_cartoon}
+\end{figure*}
+
 Each of the three lineage metrics that we discuss --- lineage length, mutation accumulation, and phenotypic volatility --- reduces a lineage to a linear sequence of states where each state represents an individual or sequence of individuals that share a common genotypic or phenotypic characteristic of interest; Figure \ref{fig:lineage_metrics_cartoon} is given as a toy example to help guide our discussion of these metrics.
@@
 
-\begin{figure*}[t]
-\includegraphics[width=\textwidth]{figs/lineage_metrics_cartoon.png}
-\caption{\small Panels (A), (B), (C), and (D) each show different ways of representing a lineage.
-The lineage has accumulated three mutations: one reverse mutation and two substitutions.
-The phenotype of ancestors changed three times.
-In (A), each ancestral state along the lineage represents a single individual.
-Here, the lineage length is equal to the number of generations spanned by the lineage (eight).
-In (B), ancestral states represent the sequence of genotypes along the lineage.
-Here, the lineage length is equal to four.
-In (C) ancestral states represent the sequence of phenotypes along the lineage.
-Here, the lineage length is equal to the number of times a different phenotype is expressed along the lineage (three).
-In (D), ancestral states are given by a particular phenotypic characteristic.
-Here, lineage length is equal to two.}
-\label{fig:lineage_metrics_cartoon}
-\end{figure*}
-
 \subsubsection{Lineage Length}
-Lineage length describes the number of ancestral states traversed by a lineage.
-If a state represents a single individual along a lineage, lineage length is a measure of the number of generations on a lineage.
-This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined, but instead determined by the life history strategies of organisms.
-In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates (\textit{e.g.} r-selected lineage) and shorter lineages imply slower replication rates (\textit{e.g.} K-selected lineage).
+Lineage length describes the number of states traversed by a lineage. If a state is defined as a single individual, %along a lineage,
+lineage length is a count of the number of generations.
+%on a lineage.
+%This form of the lineage length metric 
+Generation count is most useful in systems where generational turnover is not fixed, %algorithmically determined,
+but instead determined by the life history strategies of organisms.
+%In these systems, %lineage length can provide a proxy for replication rate where, 
+For lineages that span equal lengths of time, more generations imply faster replication rates (\textit{e.g.} r-selected lineage) while fewer generations imply slower replication rates (\textit{e.g.} K-selected lineage).
 %@ELD: I've always heard these called r- and k- selected rather than replicator, but I'm totally willing to believe these terms are used too.
@@
 
-Lineage length becomes an increasingly flexible and informative metric if we begin to consider more abstract definitions of an ancestral state along a lineage. 
-We might measure lineage length where an ancestral state represents a temporally continuous sequence of individuals that share a phenotypic or genotypic characteristic of interest.
-In these cases, lineage length only increases if the characteristic of interest changes from parent to offspring.
-For example, in an environment where organisms must perform a set of tasks to be successful, we might define ancestral state to be the set of tasks performed by an individual (\textit{e.g.} the set of logic operations performed by a digital organism in the default logic 9 Avida environment).
-In this scenario, lineage length would only increase when the set of tasks performed by an ancestor changes; any sequential ancestors that perform the identical sets of tasks would be compressed into a single state in the sequence.
+Lineage length becomes %an increasingly 
+a more flexible and informative metric if we consider more abstract definitions of %an ancestral 
+states along a lineage. 
+We might measure lineage length where a state represents a %temporally continuous 
+sequence of individuals that share a particular phenotypic or genotypic characteristic.
+%of interest. 
+In these cases, lineage length only increases when the characteristic of interest changes from parent to offspring.
+For example, in an environment where organisms must perform tasks to be successful, we might define state as the set of tasks performed by an individual.
+%(\textit{e.g.} the set of logic operations performed by a digital organism in the default logic-9 Avida environment).
+In this scenario, lineage length would only increase when the set of tasks performed by an ancestor changes; sequential ancestors that perform the identical sets of tasks would be compressed into a single state in the sequence, even if other traits differ.
 % This could be made very clear with a good example here. 
@@
 \subsubsection{Mutation Accumulation}
-Mutation accumulation defines a set of measurements that track mutational changes across a lineage.
-These changes can be measured in a variety of ways, such as the magnitude of the change (for real-valued genomes), or the total count of changes (for discrete-valued genomes).
-The type of a mutation can also be tracked, and can provide insight into the distribution of mutation types for a given lineage.
-Measures of mutation accumulation along the lineages of successful extant individuals can help to tease apart the relative importance of different types of mutational events when compared to what is expected by chance.
+Mutation accumulation defines a set of measurements that track mutational changes across a lineage. These changes can be measured %in a variety of ways, such 
+as the magnitude of the change (for real-valued genomes) or as the total count of changes (for discrete-valued genomes).
+Mutation effects can also be tracked to gain insights about their distribution along a given lineage.
+Measures of mutation accumulation along the lineages of successful individuals can help tease apart the relative importance of different types of mutational events when compared to what is expected by chance.
 
@@
 % @AML: the below points likely have many things that could be cited to support. 
-The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape, as well as how well a lineage was able to cross fitness valleys~\citep{covert_experiments_2013}.
+The number and magnitude of deleterious mutations along a lineage can tell us both about the ruggedness of the fitness landscape, and about a lineage's ability to cross fitness valleys~\citep{covert_experiments_2013}.
 Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.
@@
 
-In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be similar to the average lineage,
+In scenarios with frequent selective sweeps, the dominant lineage will likely be similar to the average lineage,
 %in the population,
-as most of the population will be closely related.
-When the population contains more phylogenetic diversity, however, it is possible for the dominant lineage to differ from the mean.
-Of course, the nature of such differences is likely informative about the evolutionary dynamics occurring in the population.
+as most of the population will be closely related. When the population contains more phylogenetic diversity, however, %it is possible for 
+the dominant lineage may differ from the mean. Of course, the nature of such differences is likely informative about the evolutionary dynamics occurring in the population. 
 %Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. 
@@
 
-These metrics operate on entire phylogenies rather than single lineages within a population.
-%Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics.
-Thus, there is no need to identify a representative organism or lineage.
-Because they %factor in 
-use date from
-the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage.
-On the other hand, because most lineages in the population share substantial history, phylogeny metrics can usually be calculated more rapidly than lineage metrics %can be calculated across
-for an entire population.
+These metrics operate on entire phylogenies rather than single lineages within a population, eliminating the need to identify a representative organism or lineage.
+Because they use data from the entire population, phylogeny metrics can be more computationally expensive to calculate than single lineage metrics.
+On the other hand, because most lineages tend to share substantial history, phylogeny metrics can usually be calculated more rapidly than full-population lineage metrics.
+%can be calculated across an entire population.
 Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc.
@@
 
-A standard technique for saving memory and time when working with phylogenties in computational systems is to ``prune" them, removing 
+A standard technique for saving memory and time when working with phylogenies in computational systems is to ``prune" them, removing 
 %any taxa that are not either currently alive or the ancestor of taxa that are alive.
@@
 
-An important distinction between phylogenies in natural versus computational systems to be aware of is that natural phylogenies are generally inferred from extant taxa, whereas computational phylogenies are directly recorded.
-The inferred phylogenies used in natural systems do not contain internal nodes except at branch points.
-They also do not contain history prior to the most recent common ancestor (MRCA) of all organisms in the extant population.
-For consistency, we will exclude pre-MRCA taxa from our analyses.
-However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.
+An important distinction between phylogenies in natural versus computational systems 
+%to be aware of 
+is that natural phylogenies are generally inferred from extant taxa, whereas computational phylogenies are directly recorded. %The inferred phylogenies used in natural systems 
+Inferred phylogenies 
+do not contain internal nodes except at branch points. They also do not contain history prior to the most recent common ancestor (MRCA) of all extant organisms. %in the extant population.
+For consistency, we exclude pre-MRCA taxa from our analyses. However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.
 
@@
 A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never change its MRCA after the very beginning of evolution %(which is informative as a way of detecting stably coexisting branches in the phylogeny).
-(which at least allows us to detect stably coexistance in the population).
+(which at least allows us to detect stably coexistence in the population).
 
@@
 
-Measurements of phylogenetic richness quantify the total amount of evolutionary history contained in a set of taxa.
-The most traditional metric of phylogenetic richness is the one simply titled ``Phylogenetic Diversity", which is calculated as the number of nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}.
+Measurements of phylogenetic richness quantify the total amount of evolutionary history contained in a set of taxa. The most traditional metric of phylogenetic richness is
+%the one simply titled
+``Phylogenetic Diversity", which is calculated as the number of nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}.
 Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}.
@@
 % @CAO: "can be calculated by calulating" ?
+% @ELD: hah, oops.
 averaging the evolutionary distinctiveness across each taxon in the population.
@@
 
-Measurements of phylogenetic regularity quantify how balanced the branches of the phylogeny are and are often the variances of values that can be calculated for individual taxa.
+Measurements of phylogenetic regularity quantify how balanced the branches are in a phylogeny and are often the variances of values calculated for individual taxa.
 Just as the mean of the pairwise distances between all taxa in the population is a measurement of phylogenetic divergence, taking their variance is a measurement of phylogenetic regularity.
@@
 \begin{figure}
-\includegraphics[width=3.5in]{figs/landscapes.png}
+\includegraphics[width=\columnwidth]{figs/landscapes.png}
+\reduceCaptionPadding
 \caption{\small The fitness landscapes used in this experiment: A) Himmelblau, B) Six-humped Camel Back, C) Shubert, and D) Composition Function 2.
@@
 % @AML: It would be neat to have a tournament size = population size treatment (pure hill climbing)
-For each test problem, we evolved populations of 1000 digital organisms under a range of mutation rates and selection strengths for 5000 generations.
-Each digital organism's genome consisted of two floating point numbers that defined its position in the fitness landscape.
+For each test problem, we evolved populations of 1000 organisms under a range of mutation rates and selection strengths for 5000 generations.
+Each organism's genome consisted of two floating point numbers that defined its position in the fitness landscape.
 We initialized populations by randomly generating a number of organisms equal to the population size. 
@@
 We evolved populations under five different tournament sizes: one, two, four, eight, and sixteen.
-Tournament size represents strength of selection where higher tournament sizes correspond to a higher strength of selection and lower tournament sizes correspond to a lower strength of selection~\citep{blickle1995mathematical}.
-A tournament size of one is equivalent to no selection pressure (\textit{i.e.} every organism in the population has an equal chance of being selected to reproduce).
+Tournament size represents strength of selection where higher tournament sizes correspond to a strong selection and lower tournament sizes correspond to weak selection~\citep{blickle1995mathematical}.
+A tournament size of one is equivalent to no selection pressure (\textit{i.e.} as every organism in the population has an equal chance of being selected to reproduce).
 Organisms selected to reproduce did so asexually.
-Each value in an offspring's genome was mutated by adding noise given by a normal distribution with a mean of 0; the `mutation rate' of a treatment defined the standard deviation used to define this normal distribution and was given as a proportion of the test problem's domain.
+Values in an offspring's genome were mutated by adding noise given by a normal distribution with a mean of 0; the `mutation rate' of a treatment defined the standard deviation used to define this normal distribution and was given as a proportion of the test problem's domain.
 We prevented mutations from causing a value to exceed the valid domain of the given problem. 
 For each problem and tournament size, we evolved populations at eight mutation rates: $1\text{e-}08$, $1\text{e-}07$, $1\text{e-}06$, $1\text{e-}05$, $1\text{e-}04$, $1\text{e-}03$, $1\text{e-}02$, and $1\text{e-}01$.
+
 % [.10, 0.01, 0.001, 0.0001, 0.00001,  0.000001, 0.0000001, 0.00000001]
+We also ran a second set of experiments to explore the impact of ecological dynamics on these metrics.
+For these experiments, we generated a stable ecology use the Eco-EA algorithm as a selection technique~\citep{goings_ecology-based_2012}.
+Eco-EA is %essentially
+a technique for creating niches that promote stable diversification in the context of an evolutionary algorithm. In %the context of 
+our test problems, we created niches associated with spatial locations across the fitness landscape. 
 For all experimental conditions, we ran ten replicates, each with a unique random number seed. 
@@
 \begin{figure}
-\includegraphics[width=3.5in]{figs/shubert_landscape.png}
+\includegraphics[width=\columnwidth]{figs/shubert_landscape.png}
+\reduceCaptionPadding
 \caption{\small A close-up on two adjacent peaks in the Shubert function fitness landscape.
@@
 Visualizing all lineages in each experimental condition in this way gave us an efficient check on our understanding of the underlying evolutionary dynamics in that condition.
-In order to make these metrics useful in the long run, it is critical that we have an accurate understanding of how various measurements correspond to the actual behavior of lineages.
-The most direct way to confirm our expectations is to visualize the path that each lineage takes across the fitness landscape, mapping the x, y, and z (fitness) coordinates of each ancestor of each member of the population~\citep{virgo_lineage_2017}.
-Creating such a visualization entails incorporating a large quantity of information into a limited space.
+In order to make these metrics useful,
+%in the long run, it is critical that 
+we must have an accurate understanding of how various measurements correspond to the actual behavior of lineages.
+The most direct way to confirm our expectations is to visualize the path that each lineage takes through the fitness landscape, mapping the x, y, and z (fitness) coordinates of each ancestor of each member of the population~\citep{virgo_lineage_2017}.
+Creating such a visualization entails condensing a large quantity of information into a limited space.
 When projected onto two dimensions, lineages can obscure parts of the fitness landscape (and each other).
@@
 
-A-Frame supports building three dimensional scenes and rendering them to a variety of platforms.
+A-Frame supports building %three dimensional 
+3D scenes and rendering them to a variety of platforms.
 In the simplest case, the visualization is rendered in WebGL, allowing it to be viewed in a standard web browser.
@@
 A-Frame also supports rendering the page with WebVR, allowing it to be viewed using various virtual reality headsets.
-These platforms allow the user to explore the data in a truly three-dimensional environment.
-Different headsets support different amounts of interaction.
+These platforms allow the user to explore the data in %a truly three-dimensional environment.
+three dimensions.
+%Different headsets support different amounts of interaction.
 For the data interpretation in this paper, we used an Oculus Rift to provide us with fine-grained control of which part of the visualization we were looking at.
@@
 % Changing mutation: here's what happens when we change mutation on each problem
+
 \begin{figure}
-\includegraphics[width=3.5in]{figs/all_mutation_rate.png}
+\includegraphics[width=\columnwidth]{figs/all_mutation_rate_crop.png}
+\reduceCaptionPadding
 \caption{\small Values of example metrics across different mutation rates for each of the four problems.
-All lineage-based metrics are calculated on the lineage of the fittest organism at the final time point (the dominant); population-level means behaved similarly.
+All lineage-based metrics are calculated on the lineage of the fittest organism at the final time point; %(the dominant);
+population-level means behaved similarly.
 All experiments shown here used a tournament size of 4.
@@
 
-Overall, our results were consistent with what evolutionary theory would predict.
+Overall, our results were consistent with %what
+evolutionary theory. %would predict. 
 As mutation rate increases, coalescence takes longer, as evidenced by the fact that the MRCA is farther back in time at higher mutation rates (see Figure \ref{fig:mutation_rate}).
-Consequently, phylogenetic richness (as measured by phylogenetic diversity) is higher at higher mutation rates.
-Phylogenetic divergence, measured here as mean pairwise distance between taxa, is similarly higher at higher mutation rates.
+Consequently, phylogenetic richness (as measured by phylogenetic diversity) is higher at high mutation rates.
+Phylogenetic divergence, measured here as mean pairwise distance between taxa, is similarly higher at high mutation rates.
 Evolutionary distinctiveness, being another measurement of phylogenetic divergence, behaved almost identically (see Supplementary Material).
-Variance of evolutionary distinctiveness and pairwise distance between taxa (phylogenetic regularity metrics) behaved very similarly to the phylogenetic divergence metrics.
+Variance of evolutionary distinctiveness and pairwise distance between taxa (phylogenetic regularity metrics) behaved similarly to the phylogenetic divergence metrics.
 This pattern makes sense, as most phylogenetic divergence on these landscapes will produce unbalanced phylogenetic trees.
 If there were stable coexistence between multiple clades, we would expect to see a reduced correlation between the phylogenetic divergence metrics and the phylogenetic regularity metrics.
-Increased mutation rate also increases the number of deleterious steps taken, a logical consequence of increasing the amount of mutation relative to the strength of selection.
+Increased mutation rate also increases the number of deleterious steps taken, a logical consequence of increasing mutation relative to strength of selection.
 
 \begin{figure}
-\includegraphics[width=3.5in]{figs/all_ts.png}
+\includegraphics[width=\columnwidth]{figs/all_ts_crop.png}
+\reduceCaptionPadding
 \caption{\small Values of example metrics across different tournament sizes for each of the four problems.
 All experiments shown here used a mutation rate of 0.001.
-All lineage-based metrics are calculated on the lineage of the fittest organism at the final time point (the dominant); population-level means behaved similarly.
+All lineage-based metrics are calculated on the lineage of the fittest organism at the final time point; population-level means behaved similarly.
 Circles are medians, vertical lines show inter-quartile range, and shaded area is a bootstrapped 95\% confidence interval around the mean.
@@
 
-Surprisingly, there is no clear effect of tournament size on the count of deleterious steps along the dominant lineage (as evidenced by the fact that the confidence intervals are all overlapping).
+\begin{figure}
+\includegraphics[width=\columnwidth]{figs/eco_mutation_rate_crop.png}
+\reduceCaptionPadding
+\caption{\small Values of example metrics across different mutation rates for each of the four problems under a diversity-preserving selection regime, Eco-EA.
+All lineage-based metrics are calculated on the lineage of the fittest organism at the final time point; population-level means behaved similarly.
+All experiments shown here used a tournament size of 4.
+Circles are medians, vertical lines show inter-quartile range, and shaded area is a bootstrapped 95\% confidence interval around the mean.
+Note that both axes are on log scales.}
+\label{fig:ecology}
+\end{figure}
+
+Surprisingly, there is no clear effect of tournament size on the count of deleterious steps along the dominant lineage (as evidenced by the fact that the confidence intervals all overlap).
 Values for all selection schemes and tournament sizes hover near 2,500, meaning that a deleterious step is taken in roughly half of the 5000 generations.
-This is partially an effect of mutation rate; at the lowest mutation rate, there is a clear trend towards fewer deleterious steps as tournament size increases (see Supplementary Material).
+This result is partially an effect of mutation rate; at the lowest mutation rate, there is a clear trend toward fewer deleterious steps as tournament sizes increase (see Supplementary Material).
 However, the effect of mutation rate on the relationship between tournament size and dominant deleterious steps is complex, particularly for composition function 2 (see Supplementary Material).
-These trends likely share a common cause with the thresholding effect evident in Figure \ref{fig:mutation_rate}, where the number of deleterious steps along the dominant lineage abruptly climbs between mutation rates of 0.0000001 and 0.00001 and remains relatively flat over other mutation rates.
-Based on an inspection of the 3-dimensional fitness landscape visualizations, we can see that this is not an effect of lineages moving from peak-to-peak; at most mutation rates, they tend to remain on a single peak.
+These trends likely share a common cause with the thresholding effect evident in Figure \ref{fig:mutation_rate}, where the number of deleterious steps along the dominant lineage abruptly climbs between mutation rates of %0.0000001 and 0.00001
+$10^{-7}$ and $10^{-5}$
+and remains relatively flat over other mutation rates. Based on an inspection of the %3-dimensional 
+3D fitness landscape visualizations, we can see that this is not an effect of lineages moving from peak-to-peak; at most mutation rates, they tend to remain on a single peak.
 Thus, we can infer that this effect is the result of a drift-like phenomenon where, at sufficiently high mutation rates, all members of the population are constantly somewhat displaced from their local fitness peak.
+
+Having %established that
+reinforced our intuition about these metrics 
+%are reasonable
+in a simple system, we can now expand them to a slightly more complex system.
+A large proportion of interesting short-term evolutionary dynamics relate to interaction between individuals in the population (i.e.
+ecological dynamics).
+In particular, such interactions often promote the stable coexistence of clades occupying different niches.
+As such, it is important to establish a baseline for how our metrics respond to ecological coexistence.
+
+Indeed, the presence of stabilizing ecological dynamics substantially changes the values we observe for most metrics (see Figure~\ref{fig:ecology}).
+Perhaps the least surprising of these is MRCA depth is far lower than it was for tournament selection, reflecting the rarity of coalescence events under these conditions.
+Consequently, phylogenetic diversity is higher, as the extant population represents a greater amount of evolutionary history.
+Relatedly, mean pairwise distance among extant taxa is higher in the presence of ecology, as clades in different niches continue to diverge.
+Interestingly, the relationship of many metrics (e.g.
+deleterious steps and phylogenetic diversity) to mutation rate is reversed in the presence of ecology.
+Explaining the underlying mechanisms behind these distinctions is beyond the scope of this paper, but the ease with which the metrics identified their presence clearly indicates their power.
 
@@
 
-The primary goal of this work is two-fold: 1) to suggest a set of metrics that will improve our capacity to quantitatively understand evolutionary histories in digital evolution experiments, and 2) to spark a conversation in the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories.
-With feedback from the community, we plan to expand the presented suite of lineage and phylogeny metrics, compiling accessible descriptions and examples of each metric.
+Our goals for this work are two-fold: 1) to suggest a set of metrics that will improve our capacity to quantitatively understand evolutionary histories in digital evolution experiments, and 2) to spark a conversation in the computational evolution community about how to quantify, interpret, and compare observed evolutionary histories.
+With feedback from the community, we will expand our suite of lineage and phylogeny metrics, compiling accessible descriptions and examples of each metric.
 
-We have demonstrated that the suite of lineage and phylogeny metrics behaves reasonably on a set of toy problems with simplistic organisms.
+We have demonstrated that %the suite of lineage and phylogeny
+these metrics behave reasonably on a set of toy problems with simple organisms.
 Having established baseline expectations for their responses to common conditions, our next step is to apply these metrics to more complex scenarios: populations of digital organisms that we evolve in a variety of qualitatively different environments where we would expect to observe a wide range of evolutionary dynamics.
@@
 
-We thank all members of the MSU Digital Evolution Lab for their helpful comments and suggestions on early drafts of this manuscript.
-This research has been supported by the National Science Foundation (NSF) BEACON Center under Cooperative Agreement DBI-0939454, by NSF Graduate Research Fellowships to ED and AL under Grant No.
-DGE-1424871, by NSF Grant No.
-DEB-1655715, which provided support for development of the software tools, and by Michigan State University through computational resources provided by the Institute for Cyber-Enabled Research and the Digital Scholarship Lab.
-Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the NSF.
-% @CAO: If needed, we can collapse a lot of the NSF grants above into a litany of grant numbers making the paragraph much shorter.
+We thank members of the MSU Digital Evolution Lab for helpful comments and suggestions on this manuscript.
+This research was supported by the National Science Foundation (NSF) through the BEACON Center (Cooperative Agreement DBI-0939454), Graduate Research Fellowships to ED and AL (Grant No.
+DGE-1424871), and NSF Grant No.
+DEB-1655715 to CO.
+Michigan State University provided computational resources through the Institute for Cyber-Enabled Research and the Digital Scholarship Lab.
+Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the NSF or MSU.
 

### 2018-04-12 source/LineageExp.h
Added paper and visualization

@@ -1,1770 +1,0 @@
-#ifndef LINEAGE_EXP_H
-#define LINEAGE_EXP_H
-
-// @includes
-#include <iostream>
-#include <string>
-#include <utility>
-#include <fstream>
-#include <sys/stat.h>
-#include <algorithm>
-#include <functional>
-
-#include "base/Ptr.h"
-#include "base/vector.h"
-#include "control/Signal.h"
-#include "Evolve/World.h"
-#include "Evolve/Resource.h"
-#include "Evolve/SystematicsAnalysis.h"
-#include "Evolve/World_output.h"
-#include "games/Othello8.h"
-#include "hardware/EventDrivenGP.h"
-#include "hardware/AvidaGP.h"
-#include "hardware/AvidaCPU_InstLib.h"
-#include "hardware/InstLib.h"
-#include "tools/BitVector.h"
-#include "tools/Random.h"
-#include "tools/random_utils.h"
-#include "tools/math.h"
-#include "tools/string_utils.h"
-
-#include "TestcaseSet.h"
-#include "OthelloHW.h"
-#include "lineage-config.h"
-
-// @constants
-constexpr int TESTCASE_FILE__DARK_ID = 1;
-constexpr int TESTCASE_FILE__LIGHT_ID = -1;
-constexpr int TESTCASE_FILE__OPEN_ID = 0;
-
-constexpr size_t SGP__TAG_WIDTH = 16;
-
-constexpr size_t REPRESENTATION_ID__AVIDAGP = 0;
-constexpr size_t REPRESENTATION_ID__SIGNALGP = 1;
-
-constexpr size_t TRAIT_ID__MOVE = 0;
-constexpr size_t TRAIT_ID__DONE = 1;
-
-constexpr size_t RUN_ID__EXP = 0;
-constexpr size_t RUN_ID__ANALYSIS = 1;
-
-constexpr size_t ANALYSIS_TYPE_ID__DEBUGGING = 0;
-
-constexpr int AGENT_VIEW__ILLEGAL_ID = -1;
-constexpr int AGENT_VIEW__OPEN_ID = 0;
-constexpr int AGENT_VIEW__SELF_ID = 1;
-constexpr int AGENT_VIEW__OPP_ID = 2;
-
-constexpr size_t SELECTION_METHOD_ID__TOURNAMENT = 0;
-constexpr size_t SELECTION_METHOD_ID__LEXICASE = 1;
[…]

### 2018-04-12 source/OthelloHW.h

@@ -1,58 +1,0 @@
-#ifndef OTHELLO_HW_H
-#define OTHELLO_HW_H
-
-#include "base/vector.h"
-#include "games/Othello8.h"
-#include "tools/random_utils.h"
-#include "tools/math.h"
-#include "tools/string_utils.h"
-
-// NOTE: we don't actually need this for test case evaluations...
-class OthelloHardware {
-
-protected:
-  using othello_t = emp::Othello8;
-  using player_t = othello_t::Player;
-  emp::vector<othello_t> dreams; ///< Let's lean into that whole 'othello dream' terminology...
-  size_t active_dream;
-  player_t playerID;
-
-public:
-  OthelloHardware(size_t dream_cnt, player_t pID=player_t::DARK)
-  : dreams(dream_cnt), active_dream(0), playerID(pID)
-  { emp_assert(dream_cnt > 0); }
-
-  othello_t & GetActiveDreamOthello() { return dreams[active_dream]; }
-
-  void SetActiveDream(size_t id) {
-    emp_assert(id < dreams.size());
-    active_dream = id;
-  }
-
-  void SetPlayerID(player_t pID) { playerID = pID; }
-  player_t GetPlayerID() const { return playerID; }
-
-  void Reset() {
-    for (size_t i = 0; i < dreams.size(); ++i) dreams[i].Reset();
-  }
-
-  void Reset(const othello_t & other) {
-    for (size_t i = 0; i < dreams.size(); ++i) {
-      dreams[i].Reset();
-      dreams[i].SetBoard(other.GetBoard());
-    }
-  }
-
-  void ResetActive() {
-    dreams[active_dream].Reset();
-  }
-
-  void ResetActive(const othello_t & other) {
-    dreams[active_dream].Reset();
-    dreams[active_dream].SetBoard(other.GetBoard());
-  }
-
-};
-
-
-#endif

### 2018-04-12 experiment/OptimizationProblemExp.h
Add experiment code

@@ -1,0 +1,816 @@
+#ifndef TOY_PROBLEM_EXP_H
+#define TOY_PROBLEM_EXP_H
+
+// @includes
+#include <iostream>
+#include <string>
+#include <utility>
+#include <fstream>
+#include <sys/stat.h>
+#include <algorithm>
+#include <functional>
+#include <ctime>
+
+#include "base/Ptr.h"
+#include "base/vector.h"
+#include "control/Signal.h"
+#include "Evolve/World.h"
+#include "Evolve/Resource.h"
+#include "Evolve/SystematicsAnalysis.h"
+#include "Evolve/World_output.h"
+#include "hardware/InstLib.h"
+#include "tools/BitVector.h"
+#include "tools/Random.h"
+#include "tools/random_utils.h"
+#include "tools/math.h"
+#include "tools/string_utils.h"
+
+#include "optimization-config.h"
+
+#include "cec2013.h"
+
+constexpr size_t RUN_ID__EXP = 0;
+constexpr size_t RUN_ID__ANALYSIS = 1;
+
+constexpr size_t DIMENSIONS = 2;
+
+constexpr size_t SELECTION_METHOD_ID__TOURNAMENT = 0;
+constexpr size_t SELECTION_METHOD_ID__LEXICASE = 1;
+constexpr size_t SELECTION_METHOD_ID__ECOEA = 2;
+constexpr size_t SELECTION_METHOD_ID__MAPELITES = 3;
+constexpr size_t SELECTION_METHOD_ID__ROULETTE = 4;
+constexpr size_t SELECTION_METHOD_ID__DRIFT = 5;
+
+const emp::vector<std::string> MUTATION_TYPES = {"normal", "normal_x", "normal_y"};
+
+// Available problems: (1-indexed in cpp library)
+// 0) F4 (2D)  --> Index: 4
+// 1) F5 (2D)  --> Index: 5
+// 2) F6 (2D)  --> Index: 6
+// 3) F7 (2D)  --> Index: 7
+// 4) F8 (2D)  --> Index: 10
+// 5) F9 (2D)  --> Index: 11
+// 6) F10 (2D) --> Index: 12
+// 7) F11 (2D) --> Index: 13
+const emp::vector<size_t> PROBLEM_MAP = {4,5,6,7,10,11,12,13};
[…]

### 2018-04-12 experiment/optimization-config.h

@@ -1,0 +1,34 @@
+#ifndef TOY_CONFIG_H
+#define TOY_CONFIG_H
+
+#include "config/config.h"
+
+EMP_BUILD_CONFIG( OptimizationConfig,
+  GROUP(DEFAULT_GROUP, "General Settings"),
+  VALUE(RUN_MODE, size_t, 0, "What mode are we running in? \n0: Native experiment\n1: Analyze mode"),
+  VALUE(RANDOM_SEED, int, -1, "Random number seed (negative value for based on time)"),
+  VALUE(POP_SIZE, size_t, 1000, "Total population size"),
+  VALUE(GENERATIONS, size_t, 5000, "How many generations should we run evolution?"),
+  VALUE(PROBLEM, size_t, 0, "Which problem?"),
+  VALUE(HINT_GRID_RES, size_t, 32, "Lexicase/eco-ea grid will be HINT_GRID_RES x HINT_GRID_RES in size."),
+  GROUP(SELECTION_GROUP, "Selection Settings"),
+  VALUE(SELECTION_METHOD, size_t, 0, "Which selection method are we using? \n0: Tournament\n1: Lexicase\n2: Eco-EA (resource)\n3: MAP-Elites\n4: Roulette\n5: Random drift"),
+  VALUE(ELITE_SELECT__ELITE_CNT, size_t, 1, "How many elites get free reproduction passes?"),
+  VALUE(TOURNAMENT_SIZE, size_t, 4, "How big are tournaments when using tournament selection or any selection method that uses tournaments?"),
+  VALUE(LEXICASE_EPSILON, double, 0.0, "Defines epsilon value for lexicase selection (score_ceil-score_floor)*THIS=epsilon"),
+  VALUE(RESOURCE_SELECT__RES_AMOUNT, double, 50.0, "Initial resource amount (for all resources)"),
+  VALUE(RESOURCE_SELECT__RES_INFLOW, double, 50.0, "Resource in-flow (amount)"),
+  VALUE(RESOURCE_SELECT__OUTFLOW, double, 0.05, "Resource out-flow (rate)"),
+  VALUE(RESOURCE_SELECT__FRAC, double, 0.0025, "Fraction of resource consumed."),
+  VALUE(RESOURCE_SELECT__MAX_BONUS, double, 5.0, "What's the max bonus someone can get for consuming a resource?"),
+  VALUE(RESOURCE_SELECT__COST, double, 0.0, "Cost of using a resource?"),
[…]

### 2018-04-12 experiment/optimization_problems.cc

@@ -1,0 +1,35 @@
+// This is the main function for the NATIVE version of this project.
+
+#include <iostream>
+
+#include "base/vector.h"
+
+#include "config/command_line.h"
+#include "config/ArgManager.h"
+
+#include "optimization-config.h"
+#include "OptimizationProblemExp.h"
+
+int main(int argc, char* argv[])
+{
+  // Read configs.
+  std::string config_fname = "configs.cfg";
+  auto args = emp::cl::ArgManager(argc, argv);
+  OptimizationConfig config;
+  config.Read(config_fname);
+
+  if (args.ProcessConfigOptions(config, std::cout, config_fname, "../optimization-config.h") == false)
+    exit(0);
+  if (args.TestUnknown() == false)
+    exit(0);
+
+  std::cout << "==============================" << std::endl;
+  std::cout << "|    How am I configured?    |" << std::endl;
+  std::cout << "==============================" << std::endl;
+  config.Write(std::cout);
+  std::cout << "==============================\n"
+            << std::endl;
+
+  OptimizationProblemExp e(config);
+  e.Run();
+}

### 2018-04-12 experiment/scripts/potHoles.py

@@ -1,0 +1,120 @@
+import argparse, os, errno
+
+'''
+potHoles.py
+Given a set of job_not_done logs (example log contents below), rebuild qsub file to submit jobs. 
+
+== Example contents of job_not_done log: ==
+P0_G32_M0.0001_S1_E1_LE1e-09_1783/run.log
+P0_G32_M0.0001_S1_E1_LE1e-12_1796/run.log
+P0_G32_M0.0001_S5_E0_6431/run.log
+P0_G32_M0.0001_S5_E0_6433/run.log
+P0_G32_M0.0001_S5_E0_6437/run.log
+
+'''
+
+qsub_base = '''#!/bin/bash -login
+
+### Configure job:
+#PBS -l walltime=04:00:00:00
+#PBS -l feature=intel16
+#PBS -l mem=8gb
+#PBS -t [[JOB_CONFIG:-t]]
+#PBS -N POTHOLES
+
+### load necessary modules, e.g.
+module load powertools
+
+# General Parameters.
+EXEC=command.sh
+
+'''
+
+def main():
+    parser = argparse.ArgumentParser(description="Pot hole filler.")
+    parser.add_argument("-d", "--data_directory", type=str, action="append", help="Target experiment directory.")
+    parser.add_argument("-l", "--not_done_log", type=str, action="append", help="Location of log of unfinished jobs.")
+    parser.add_argument("-c", "--condition", type=str, action="append", help="At least one given condition string must be job name to include in resubs.")
+
+    args = parser.parse_args()
+
+    data_directories = args.data_directory
+    not_done_logs = args.not_done_log
+    filter_jobs = bool(args.condition)
+    include_conditions = args.condition
+    
+    def JobFilter(job_name):
+        if (not filter_jobs): return True
+        for condition in include_conditions:
+            if (condition in job_name):
+                return True
+        return False
+
+    print("Data directories: {}".format(data_directories))
+    print("Not done logs: {}".format(not_done_logs))
+
+    if (filter_jobs):
+        print("Filtering jobs on conditions: {}".format(include_conditions))
+
+    if (not data_directories):
+        print("Must provide at least one data directory!")
+        exit()
+    if (not not_done_logs):
+        print("Must provide at least one log of unfinished jobs!")
+        exit()
+    
+    unfinished_jobs = []
+    # Aggregate unfinished jobs.
[…]

### 2018-04-12 analysis/cec_python_library/analyze_landscapes.py
Add script to generate heatmaps

@@ -1,0 +1,52 @@
+#!/home/emily/anaconda2/bin/python
+###############################################################################
+# Version: 1.1
+# Last modified on: 3 April, 2016 
+# Developers: Michael G. Epitropakis
+#      email: m_(DOT)_epitropakis_(AT)_lancaster_(DOT)_ac_(DOT)_uk 
+###############################################################################
+from cec2013.cec2013 import *
+import numpy as np
+import sys
+import pandas as pd
+import matplotlib.pyplot as plt
+
+def main():
+
+	problem = int(sys.argv[1])
+
+	# Create function
+	f = CEC2013(problem)
+
+	lx = f.get_lbound(0)
+	ly = f.get_lbound(1)
+	ux = f.get_ubound(0)
+	uy = f.get_ubound(1)
+
+	xs = np.linspace(lx, ux, 1000)
+	ys = np.linspace(ly, uy, 1000)
+	xs, ys = np.meshgrid(xs, ys)
+	# print(lx, ly, ux, uy, xs, ys)	
+	# zs = np.array([f.evaluate([x, y]) for x in xs for y in ys])
+	zs = np.array([f.evaluate([x,y]) for x,y in zip(np.ravel(xs), np.ravel(ys))])
+	# print(zs)
+	zs = zs.reshape(xs.shape)
+	# zs = zs.reshape(len(xs), len(ys))
+	# print(zs)
+	plt.imsave("fun_"+ str(problem) +".jpg", zs, format="jpg", cmap="gray")
+	
+	with open("fun_"+str(problem)+"_bounds.csv", "w") as outfile:
+		outfile.write(" ".join(["x:", str(lx), str(ux)]) + "\n")
+		outfile.write(" ".join(["y:", str(ly), str(uy)]) + "\n")
+		outfile.write(" ".join(["z:", str(zs.min()), str(zs.max())]) + "\n")
+
+	#print(df)
+	# Evaluate :-)
+	# x = np.ones(2)
+	# value = f.evaluate(x)
+
+
+
+
+if __name__ == "__main__":
+	main()

### 2018-04-12 analysis/cec_python_library/demo.py

@@ -1,57 +1,0 @@
-#!/usr/bin/env python
-###############################################################################
-# Version: 1.1
-# Last modified on: 3 April, 2016 
-# Developers: Michael G. Epitropakis
-#      email: m_(DOT)_epitropakis_(AT)_lancaster_(DOT)_ac_(DOT)_uk 
-###############################################################################
-from cec2013.cec2013 import *
-import numpy as np
-
-def main():
-	print 70*"="
-	# Demonstration of all functions
-	for i in range(1,21):
-		# Create function
-		f = CEC2013(i)
-
-		# Create position vectors
-		x = np.ones( f.get_dimension() )
-
-		# Evaluate :-)
-		value = f.evaluate(x)
-		print "f", i, "(", x, ") = ", f.evaluate(x)
-
-	print 70*"="
-	# Demonstration of using how_many_goptima function
-	for i in range(1,21):
-		# Create function
-		f = CEC2013(i)
-		dim = f.get_dimension()
-
-		# Create population of position vectors
-		pop_size = 10
-		X = np.zeros( (pop_size, dim) )
-		ub =np.zeros( dim )
-		lb =np.zeros( dim )
-		# Get lower, upper bounds
-		for k in range(dim):
-			ub[k] = f.get_ubound(k)
-			lb[k] = f.get_lbound(k)
-
-		# Create population within bounds
-		fitness = np.zeros( pop_size )
-		for j in range(pop_size):
-			X[j] = lb + (ub - lb) * np.random.rand( 1 , dim )
-			fitness[j] = f.evaluate(X[j])
-
-		# Calculate how many global optima are in the population
-		accuracy = 0.001
-		count, seeds = how_many_goptima(X, f, accuracy)
-		print "In the current population there exist", count, "global optimizers."
-		print "Global optimizers:", seeds
-
-	print 70*"="
-
-if __name__ == "__main__":
-	main()

### 2018-04-12 analysis/cec_python_library/extract_dominant_lineaage_info.py
Updated dominant lineage info

@@ -1,16 +1,20 @@
 #!/home/emily/anaconda2/bin/python
-###############################################################################
-# Version: 1.1
-# Last modified on: 3 April, 2016 
-# Developers: Michael G. Epitropakis
-#      email: m_(DOT)_epitropakis_(AT)_lancaster_(DOT)_ac_(DOT)_uk 
-###############################################################################
+
 from cec2013.cec2013 import *
 import numpy as np
 import sys
 import pandas as pd
 import glob
 import os
+
+# This script expects two command line arguments:
+#   - a glob-style pattern indicating which directories to analyze
+#         (remember to put quotation marks around it)
+#   - the name of a file to store data in
+
+# It will extract the path (x, y, and z coordinates) taken by the fittest
+# lineage from the phylogeny_5000.csv file from each directory matched by
+# the glob pattern provided in the first command-line argument.
 
 def main():
 
@@ -77,7 +81,7 @@ def main():
 			# print(df.loc[curr_id, :])
 			fits.append(df.loc[curr_id, "fitness"])
 			
-			# path.append(" ".join(["{:.2f}".format(i) for i in [df.loc[curr_id, "x"], df.loc[curr_id, "y"], df.loc[curr_id, "fitness"]]]))
+			 path.append(" ".join(["{:.2f}".format(i) for i in [df.loc[curr_id, "x"], df.loc[curr_id, "y"], df.loc[curr_id, "fitness"]]]))
 			if (df.loc[curr_id, "parent_id"] == 1):
 				break
 			x_magnitude += abs(df.loc[df.loc[curr_id, "parent_id"], "x"] - df.loc[curr_id, "x"]) 

### 2018-04-12 analysis/cec_python_library/extract_lineage_fitness.py

@@ -1,42 +1,0 @@
-#!/home/emily/anaconda2/bin/python
-###############################################################################
-# Version: 1.1
-# Last modified on: 3 April, 2016 
-# Developers: Michael G. Epitropakis
-#      email: m_(DOT)_epitropakis_(AT)_lancaster_(DOT)_ac_(DOT)_uk 
-###############################################################################
-from cec2013.cec2013 import *
-import numpy as np
-import sys
-import pandas as pd
-
-def main():
-
-	problem = int(sys.argv[1])
-	filename = sys.argv[2]
-
-	# Create function
-	f = CEC2013(problem)
-	assert(f.get_dimension() == 2)
-
-	df = pd.read_csv(filename)
-	df.rename(str.strip, axis="columns", inplace=True)
-	# print(df.columns.values)
-	genotypes = [i.strip("[] ").split() for i in df["info"]]
-	genotypes = map(lambda x: [float(x[0]), float(x[1])], genotypes)
-	df["fitness"] = [f.evaluate(np.array(x)) for x in genotypes]
-	genotypes = zip(*genotypes)
-
-	df["x"] = genotypes[0]
-	df["y"] = genotypes[1]
-
-	print(df)
-	# Evaluate :-)
-	# x = np.ones(2)
-	# value = f.evaluate(x)
-
-
-
-
-if __name__ == "__main__":
-	main()

### 2018-04-13 analysis/data_analysis.Rmd
Add analysis

--- 2018-04-02 analysis/data_analysis.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
+
+# Load libraries
 library(ggplot2)
@@
 
-```{r, echo=FALSE}
-alex_real_value_data <- read_csv("alex_real_value_data.csv")
-emily_real_value_data <- read_csv("real_value_data.csv")
-real_value_data <- merge(alex_real_value_data, emily_real_value_data, all = TRUE)
+This document contains all of the R code used to analyze data for this paper. The input csv files were generated from the raw data via Python scripts that are also included in this directory.
+
+First let's read in the data:
+
+```{r, message=FALSE, warning=FALSE}
+
+# Pull in data over all evolutionary time
+real_value_data  <- read.csv("real_value_data.csv")
+
+# Convert problem and selection type codes to informative names. Convert things that should be factors to factors
 real_value_data$problem <- sapply(as.character(real_value_data$p),switch,"0"="Himmelblau","1"="Six-hump camel back","2"="Shubert", "6"="Composition function 2")
@@
 real_value_data$selection <- sapply(as.character(real_value_data$selection),switch,"0"="Tournament","1"="Lexicase","2"="Eco-EA", "3"="MAP-Elites", "4"="Roulette", "5"="Drift")
+real_value_data$ts <- with(real_value_data, ifelse(selection=="Drift", 1, ts)) # Drift is equivalent to tournament size 1
 real_value_data$selection <- as.factor(real_value_data$selection)
 real_value_data$seed <- as.factor(real_value_data$seed)
-#real_value_data$ts <- as.factor(real_value_data$ts)
 real_value_data$elite <- as.factor(real_value_data$elite)
-#real_value_data$mutation_rate <- as.factor(real_value_data$mutation_rate)
-tournament_data <- subset(real_value_data, real_value_data$selection == "Tournament" & real_value_data$elite == 0)
+
+# Grab the portion of data we're working with
+tournament_data <- subset(real_value_data, (real_value_data$selection == "Tournament" | real_value_data$selection=="Drift") & real_value_data$elite == 0)
 tournament_endpoints <- filter(tournament_data, update==5000)
+tournament_endpoints$disp_pairwise_distance <- tournament_endpoints$variance_pairwise_distance/tournament_endpoints$mean_pairwise_distance
+tournament_endpoints$disp_evo_distinct <- tournament_endpoints$variance_evolutionary_distinctiveness/tournament_endpoints$mean_evolutionary_distinctiveness
 
-all_dom_data <- read_csv("~/repos/ALife2018-Lineage/analysis/cec_python_library/all_dom_data.csv")
+# Pull in data that was calculated post-hoc just for the end points
+all_dom_data <- read_csv("all_dom_data.csv")
 all_dom_data$update <- 5000
@@
 
+# Combine data
 tournament_endpoints <- merge(tournament_endpoints, all_dom_data)
+
+# Melt the data so that we can easily make graphs comparing different variables
 melted_tournament_data <- melt(tournament_endpoints, id.vars = c("update", "selection", "seed", "elite", "mutation_rate", "le", "p", "problem", "g", "ts", "TOURNAMENT_SIZE", "FITNESS_INTERVAL", "RESOURCE_SELECT__OUTFLOW", "MUTATION_STD", "SELECTION_METHOD", "RESOURCE_SELECT__RES_AMOUNT", "HINT_GRID_RES", "RESOURCE_SELECT__RES_INFLOW", "LEXICASE_EPSILON", "SYSTEMATICS_INTERVAL", "RESOURCE_SELECT__FRAC", "RESOURCE_SELECT__COST", "RESOURCE_SELECT__MAX_BONUS", "POP_SNAPSHOT_INTERVAL", "POP_SIZE", "ELITE_SELECT__ELITE_CNT", "GENERATIONS"))
-end_points_vars_of_interest <- melted_tournament_data %>%  filter(variable %in% c("mrca_depth", "dominant_deleterious_steps", "mean_evolutionary_distinctiveness", "variance_evolutionary_distinctiveness", "current_phylogenetic_diversity", "mean_pairwise_distance", "variance_pairwise_distance", "phenotypic_volatility"))
 
-labels <- c(current_phylogenetic_diversity = "Phylogenetic diversity", dominant_deleterious_steps = "Dominant deleterious steps", mean_evolutionary_distinctiveness="Mean evo.
+# Pull out just the variables we care about
+end_points_vars_of_interest <- melted_tournament_data %>%  filter(variable %in% c("mrca_depth", "dominant_deleterious_steps", "current_phylogenetic_diversity", "mean_pairwise_distance", "phenotypic_volatility", "total_magnitude"))
+
+# Pull out vairables for supplementary info
+extended_end_points_vars_of_interest <- melted_tournament_data %>%  filter(variable %in% c("mrca_depth", "dominant_deleterious_steps", "current_phylogenetic_diversity", "mean_pairwise_distance", "phenotypic_volatility", "total_magnitude", "variance_pairwise_distance", "mean_evolutionary_distinctiveness", "variance_evolutionary_distinctiveness"))
+
+# Convert variable names to useful axis labels
+labels <- c(current_phylogenetic_diversity = "Phylogenetic diversity", dominant_deleterious_steps = "Deleterious steps", mean_evolutionary_distinctiveness="Mean evo.
 distinctiveness", mean_pairwise_distance="Mean pairwise distance", mrca_depth="MRCA depth", variance_evolutionary_distinctiveness="Variance evo.
-distinctiveness", variance_pairwise_distance = "Variance pairwise distance", phenotypic_volatility="Phenotypic volatility")
+distinctiveness", variance_pairwise_distance = "Variance pairwise distance", phenotypic_volatility="Phenotypic volatility", total_magnitude="Mutation accumulation")
 
@@
 
-```{r}
-ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$mutation_rate==.001), aes(x=ts, y=value, color=problem, group=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="line", fun.y=median)+ theme_classic() + scale_x_log10("Tournament size", breaks=c(2,4,8,16)) + scale_y_continuous("", labels=comma) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels)) + scale_color_discrete("Problem") + theme(legend.position = "bottom")
-ggsave("../figs/all_ts.png", width=7, height=5) 
+## Tournament size results
+
+Okay, now that we've got out data, let's plot it. First, let's look at how tournament size effects each of our variables (holding mutation rate constant at .001).
+
+```{r, message=FALSE, warning=FALSE}
+ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$mutation_rate==.001), aes(x=ts, y=value, fill=problem, color=problem, group=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Tournament size", breaks=c(1,2,4,8,16)) + scale_y_log10("", labels=scientific, breaks = scales::trans_breaks("log10", function(x) 10^x)) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels),ncol=2) + scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+ggsave("../figs/all_ts.png", width=3.5, height=4.5) 
 ```
 
-```{r}
-ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$ts==4), aes(x=mutation_rate, y=value, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="line", fun.y=median)+ theme_classic() + scale_x_log10("Mutation rate", labels=scientific) + scale_y_continuous("", labels=comma) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels)) +scale_color_discrete("Problem") +theme(legend.position = "bottom")
-ggsave("../figs/all_mutation_rate.png", width=7, height=5)
+We don't have space for this in the paper, but let's also take a look at the next mutation rate up, just to make sure this isn't being driven by the mutation rate.
+While we're at it, lets also graph some additional metrics:
+
+```{r, message=FALSE, warning=FALSE}
+ggplot(subset(extended_end_points_vars_of_interest, end_points_vars_of_interest$mutation_rate==.01), aes(x=ts, y=value, fill=problem, color=problem, group=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Tournament size", breaks=c(1,2,4,8,16)) + scale_y_log10("", labels=scientific, breaks = scales::trans_breaks("log10", function(x) 10^x)) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels),ncol=2) + scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+```
+
+And lets look at the lowest mutation rate, too:
+
+```{r, message=FALSE, warning=FALSE}
+ggplot(subset(extended_end_points_vars_of_interest, end_points_vars_of_interest$mutation_rate==.00000001), aes(x=ts, y=value, fill=problem, color=problem, group=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Tournament size", breaks=c(1,2,4,8,16)) + scale_y_log10("", labels=scientific, breaks = scales::trans_breaks("log10", function(x) 10^x)) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels),ncol=2) + scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+```
+
+## Mutation rate results
+
+Okay, now holding tournament size constant at 4, what happens when we vary mutation rate?
+
+```{r, message=FALSE, warning=FALSE}
+ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$ts==4), aes(x=mutation_rate, y=value, fill=problem, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Mutation rate", labels=scientific) + scale_y_log10("", breaks = scales::trans_breaks("log10", function(x) 10^x), labels=scientific) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels), ncol=2) +scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+ggsave("../figs/all_mutation_rate.png", width=3.5, height=4.5)
 
 ```
+
+And lets try a couple moure tournament sizes, just to be sure:
+
+```{r, message=FALSE, warning=FALSE}
+ggplot(subset(extended_end_points_vars_of_interest, end_points_vars_of_interest$selection=="Drift"), aes(x=mutation_rate, y=value, fill=problem, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Mutation rate", labels=scientific) + scale_y_log10("", breaks = scales::trans_breaks("log10", function(x) 10^x), labels=scientific) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels), ncol=2) +scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+
+```
+
+```{r, message=FALSE, warning=FALSE}
+ggplot(subset(extended_end_points_vars_of_interest, end_points_vars_of_interest$ts==2), aes(x=mutation_rate, y=value, fill=problem, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Mutation rate", labels=scientific) + scale_y_log10("", breaks = scales::trans_breaks("log10", function(x) 10^x), labels=scientific) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels), ncol=2) +scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+
+```
+
+```{r, message=FALSE, warning=FALSE}
+ggplot(subset(extended_end_points_vars_of_interest, end_points_vars_of_interest$ts==8), aes(x=mutation_rate, y=value, fill=problem, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Mutation rate", labels=scientific) + scale_y_log10("", breaks = scales::trans_breaks("log10", function(x) 10^x), labels=scientific) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels), ncol=2) +scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+
+```
+
+## Coexistance results
+
+A lot of these metrics will be substantially changed by having coexistence. Let's try using Eco-EA to generate coexistence and see how it affects our metrics.
+
+```{r, message=FALSE, warning=FALSE}
+#Coexistence figure
+
+# Grab the portion of data we're working with
+eco_ea_data <- subset(real_value_data, real_value_data$selection == "Eco-EA" & real_value_data$elite == 1)
+eco_ea_endpoints <- filter(eco_ea_data, update==5000)
+eco_ea_endpoints$disp_pairwise_distance <- eco_ea_endpoints$variance_pairwise_distance/eco_ea_endpoints$mean_pairwise_distance
+eco_ea_endpoints$disp_evo_distinct <- eco_ea_endpoints$variance_evolutionary_distinctiveness/eco_ea_endpoints$mean_evolutionary_distinctiveness
+
+eco_ea_endpoints <- merge(eco_ea_endpoints, all_dom_data)
+melted_eco_ea_data <- melt(eco_ea_endpoints, id.vars = c("update", "selection", "seed", "elite", "mutation_rate", "le", "p", "problem", "g", "ts", "TOURNAMENT_SIZE", "FITNESS_INTERVAL", "RESOURCE_SELECT__OUTFLOW", "MUTATION_STD", "SELECTION_METHOD", "RESOURCE_SELECT__RES_AMOUNT", "HINT_GRID_RES", "RESOURCE_SELECT__RES_INFLOW", "LEXICASE_EPSILON", "SYSTEMATICS_INTERVAL", "RESOURCE_SELECT__FRAC", "RESOURCE_SELECT__COST", "RESOURCE_SELECT__MAX_BONUS", "POP_SNAPSHOT_INTERVAL", "POP_SIZE", "ELITE_SELECT__ELITE_CNT", "GENERATIONS"))
+eco_ea_vars_of_interest <- melted_eco_ea_data %>%  filter(variable %in% c("mrca_depth", "dominant_deleterious_steps", "current_phylogenetic_diversity", "mean_pairwise_distance", "phenotypic_volatility", "total_magnitude"))
+
+extended_eco_ea_vars_of_interest <- melted_eco_ea_data %>%  filter(variable %in% c("mrca_depth", "dominant_deleterious_steps", "current_phylogenetic_diversity", "mean_pairwise_distance", "phenotypic_volatility", "total_magnitude", "mean_evolutionary_distinctiveness", "variance_evolutionary_distinctiveness", "variance_pairwise_distance"))
+
+ggplot(eco_ea_vars_of_interest, aes(x=mutation_rate, y=value, fill=problem, color=problem, group=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Mutation rate") + scale_y_log10("", labels=scientific, breaks = scales::trans_breaks("log10", function(x) 10^x)) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels),ncol=2) + scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+ggsave("../figs/eco_mutation_rate.png", width=3.5, height=4.5)
+
+```
+
+Looks like theres some interesting stuff going on with some of the other variables, but we definitely don't have space in the paper to explore it:
+
+```{r, message=FALSE, warning=FALSE}
+ggplot(extended_eco_ea_vars_of_interest, aes(x=mutation_rate, y=value, fill=problem, color=problem, group=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="smooth", fun.data="mean_cl_boot")+ theme_classic() + scale_x_log10("Mutation rate") + scale_y_log10("", labels=scientific, breaks = scales::trans_breaks("log10", function(x) 10^x)) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels),ncol=2) + scale_color_discrete("Problem") + theme(legend.position = "bottom", legend.title = element_text(size=10), legend.text = element_text(size=6), strip.text = element_text(size=6), axis.text = element_text(size=6)) + guides(color=guide_legend(ncol=2), fill="none")
+```
